---
title: "06: Emerging Hot Spot Analysis"
author: "Chia Ka Ying"
date: "2026-09-26"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

Emerging Hot Spots Analysis (EHSA) is a spatio-temporal statistical technique that identifies and categorises areas where clusters of high or low values of a phenomenon are developing, intensifying or diminishing over time. It combines the Getis-Ord Gi* statistic, which measures spatial clustering, with the Mann-Kendall trend test, a non-parametric trend analysis, to detect trends in these clusters within a “space-time cube” of data. This process allows for the detection of patterns such as new hot spots, persistent trends, or diminishing cold spots across different confidence levels.

The analysis consist of four main steps:

- Building a space-time cube.
- Calculating Getis-Ord local Gi* statistic for each bin by using an FDR correction.
- Evaluating these hot and cold spot trends by using Mann-Kendall trend test.
- Categorising each study area location by referring to the resultant trend z-score and p-value for each location with data, and with the hot spot z-score and p-value for each bin.

## 1. Getting started

### 1.1 Installing and Loading the R Packages

```{r}
pacman::p_load(sf, sfdep, tmap, 
               plotly, tidyverse, 
               Kendall)
```

### 1.2 Data

I will now import `hunan` and `Hunan_GDPPC`, which are a shapefile and csv respectively.

#### 1.2.1 Importing geospatial data

```{r}
hunan <- st_read(
  dsn = "data/geospatial", 
  layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

```{r}
GDPPC <- read_csv(
  "data/aspatial/Hunan_GDPPC.csv")
```

## 2. Creating a Time Series Cube

*spacetime()* of sfdep is used to create an spatio-temporal cube, which is a spacetime object class. There are four important aspects of the spacetime class:

- the data, which is a data.frame object (`Hunan_GDPPC`).
- the geometry, which is an sf object (`hunan`).
- the location identifiers from both the data.frame object and sf object (county from `hunan` and `Hunan_GDPPC`).
- the time column from the data.frame object (year from `Hunan_GDPPC`).

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

Next, *is_spacetime_cube()* of sfdep package will be used to verify if `GDPPC_st` is indeed a spacetime cube object.

```{r}
is_spacetime_cube(GDPPC_st)
```

The TRUE return confirms that `GDPPC_st` object is indeed a spacetime cube.

## 3. Computing Gi*

### 3.1 Deriving the spatial weights

I will first compute to identify neighbors and to derive an inverse distance weights:

- *activate()* of dplyr package is used to activate the geometry context. This ensures that it does not count for every year, i.e., it will have GDPPC for each year but the nb and wt will be the same for every year.
- *mutate()* of dplyr package is used to create two new columns nb and wt.
- Then we will activate the data context again and copy over the nb and wt columns to each time-slice using *set_nbs()* and *set_wts()*.
- Row order is very important so do not rearrange the observations after using *set_nbs()* or *set_wts()*.
- Take note to indicate *nb = include_self()* as this is using Gi*.

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb, 
                             geometry, 
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

### 3.2 Calculate the local Gi* for each location

We can use these new columns to manually calculate the local Gi* for each location. We can do this by grouping by Year and using *local_gstar_perm()* of sfdep package. After which, we use *unnest()* to unnest `gi_star` column of the newly created `gi_stars` data.frame to ensure that it will have the calculation for every year.

```{r}
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

## 4. Mann-Kendall Test

A monotonic series or function is one that only increases or decreases, and never changes direction. So long as the function either stays flat or continues to increase, it is monotonic. Here are the null and alternative hypotheses.

- H0: No monotonic trend
- H1: Monotonic trend is present

Here is the interpretation we can draw:

- Reject the null-hypothesis null if the p-value is smaller than the alpha value (i.e., 1-confident level)
- Tau ranges between -1 and 1 where:
  - 1 is a perfectly decreasing series; and
  - 1 is a perfectly increasing series.

### 4.1 Mann-Kendall Test on Gi

With these Gi* measures we can then evaluate each location for a trend using the Mann-Kendall test. This will be done for Changsha county.

```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

### 4.2 Plotting Mann-Kendall Test

Next, we plot the result by using ggplot2 functions.

```{r}
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

### 4.3 Interactive Mann-Kendall Plot

An interactive plot can be done using *ggplotly()* of plotly package.

```{r}
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

### 4.4 Printing Mann-Kendall test report

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

In the above result, sl is the p-value. With reference to the results, we will reject the null hypothesis and infer that there is a slight upward trend.

### 4.5 Mann-Kendall test data.frame

We can replicate this for each location by using *group_by()* of dplyr package.

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
head(ehsa)
```

We can also sort to show significant emerging hot/cold spots.

```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:10)
head(emerging)
```

## 5. Performing Emerging Hotspot Analysis

Lastly, we will perform EHSA analysis by using *emerging_hotspot_analysis()* of sfdep package. It takes a spacetime object x (i.e., `GDPPC_st`), and the quoted name of the variable of interest (i.e. `GDPPC`) for *.var* argument. The k argument is used to specify the number of time lags which is set to 1 by default. Lastly, nsim map numbers of simulation to be performed.

```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

### 5.1 Visualising the distribution of EHSA classes

*ggplot2* functions is used to reveal the distribution of EHSA classes as a bar chart.

```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

Figure above shows that sporadic cold spots class has the highest number of county.

### 5.2 Visualising EHSA

I will now visualise the geographic distribution EHSA classes. This would requiring joining `hunan` and `ehsa` first.

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

tmap functions will be used to plot a categorical choropleth map.

```{r}
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(fill_alpha = 0.4)
```