---
title: "Hands-On_Ex01"
---

## 1. Getting Started: Loading R Packages

Using the p_load function of pacman package, I will load sf and tidyverse R packages into the RStudio environment.

```{r}
pacman::p_load(sf, tidyverse)
```

<!-- ## Importing Data -->

## 2. Importing Geospatial Data

Using the st_read() of sf package, I will be loading the following three geospatial data and converting them to the desired formats:

- `MP14_SUBZONE_WEB_PL`, a polygon feature layer in GeoJSON format, which will be converted to an ESRI shapefile format;

- `PreSchool`, point feature layer in GeoJSON format, which will be converted to a KML file format; and

- `CyclingPath`, a line feature layer in ESRI shapefile format.

### 2.1 **Importing polygon feature data and coverting to shapefile format**

Using the st_read() function of sf package, I will be importing the polygon feature layer in GeoJSON format file `MP14_SUBZONE_WEB_PL` into R. I will also be converting it from a GeoJSON format to an ESRI shapefile format using the st_write() function. To ensure that any existing file is not overwritten, I have included a delete_dsn = FALSE argument and #\| eval: false chunk option to display the code in the final rendered code but prevents RStudio from executing it during the rendering process to prevent an error.

```{r}
#| eval: false
mpsz <- st_read("data/geospatial/MP14_SUBZONE_NO_SEA_PL.geojson")

dir.create("data/geospatial/mpsz", recursive = TRUE, showWarnings = FALSE) #create a folder

st_write(mpsz, "MP14_SUBZONE_NO_SEA_PL.shp", driver = "ESRI Shapefile", delete_dsn = TRUE)
```

The message above reveals that the geospatial objects are multipolygon features. There are a total of 323 multipolygon features and 13 fields in `mpsz` simple feature data frame. `mpsz` is in **WGS 84**, which is a geodetic system and thusgeographic (3D/spherical) rather than projected (2D/flat). The bounding box provides the x extend and y extend of the data.

### 2.2 **Importing** point feature layer **and coverting to KML format**
