---
title: "Hands-On Exercise 01"
---

## 1. Getting Started: Loading R Packages

Using the *p_load* function of pacman package, I will load sf and tidyverse R packages into the R environment.

```{r}
pacman::p_load(sf, tidyverse)
```

## 2. Importing Geospatial Data

Using the *st_read()* of sf package, I will be loading the following three geospatial data and converting them to the desired formats:

- `MP14_SUBZONE_NO_SEA_PL`, a polygon feature layer in GeoJSON format, which will be converted to an ESRI shapefile format;

- `PreSchoolsLocation`, point feature layer in GeoJSON format, which will be converted to a KML file format; and

- `CyclingPath`, a line feature layer in ESRI shapefile format.

### 2.1 Importing polygon feature data and coverting to shapefile format (MP14_SUBZONE_NO_SEA_PL)

Using the *st_read()* function of sf package, I will be importing the polygon feature layer in **GeoJSON** format file `MP14_SUBZONE_WEB_PL` into R. I will also be converting it from a **GeoJSON** format to an **ESRI shapefile** format using the *st_write()* function. To ensure that any existing file is not overwritten, I have included a *delete_dsn = FALSE* argument and *#\| eval: false* chunk option to display the code in the final rendered code but prevents RStudio from executing it during the rendering process to prevent an error.

```{r}
mpsz <- st_read("data/geospatial/MP14_SUBZONE_NO_SEA_PL.geojson")
```

```{r}
#| eval: false #prevent error when rendering

dir.create("data/geospatial/mpsz", recursive = TRUE, showWarnings = FALSE) #create folder

st_write(mpsz, "data/geospatial/mpsz/MP14_SUBZONE_NO_SEA_PL.shp", 
         driver = "ESRI Shapefile", delete_dsn = FALSE) #write files in ESRI format and prevent overwrite
```

The message above reveals that the geospatial objects are multipolygon features. There are a total of 323 multipolygon features and 13 fields in `mpsz` simple feature data frame. `mpsz` is in **WGS 84**, which is a geodetic system and thusgeographic (3D/spherical) rather than projected (2D/flat). The bounding box provides the x extend and y extend of the data.

### 2.2 Importing point feature layer and coverting to KML format (PreSchoolsLocation)

Similary, using the *st_read()* function of sf package, I will be importing the point feature layer in **GeoJSON** format file `PreSchoolsLocation` into R. I will also be converting it from a **GeoJSON** format to **KML** format using the *st_write()* function. To ensure that any existing file is not overwritten, I have included a *delete_dsn = FALSE* argument and *#\| eval: false* chunk option to display the code in the final rendered code but prevents RStudio from executing it during the rendering process to prevent an error.

```{r}
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")
```

```{r}
#| eval: false #prevent error when rendering

dir.create("data/geospatial/preschool", recursive = TRUE, showWarnings = FALSE) #create folder

st_write(preschool, "data/geospatial/preschool/preschool.kml", 
         driver = "KML", delete_dsn = FALSE) #write files and prevent overwrite
```

The message above reveals that `preschool` is a point feature data frame. There are a total of 2290 features and 2 fields. `preschool` is in **WGS 84**, which is the same as `mpsz`.

### 2.3 Importing polyline feature data in shapefile format (CyclingPath)

The code chunk below uses *st_read()* function of sf package to import CyclingPath shapefile into R as a line feature data frame.

```{r}
cyclingpath = st_read(dsn = "data/geospatial/CyclingPath", 
                         layer = "CyclingPath")
```

There are a total of 4830 features and 7 fields in cyclingpath sf data frame. The geospatial entities are capture in multilinestring object. Unlike the earlier two data sets, this data set is in **SVY21** projected coordinates system.

## 3. Checking the Content of a Simple Feature Data Frame

Before performing data wrangling, I will be learning more about the `mpsz` data set using *st_geometry(),* *glimpse()* and *head().*

### 3.1 Working with st_geometry

To retrieve the geometry list-column, I will be using the below code chunk, which is the more general way, rather than *mpsz\$geom* or *mpsz\[\[1\]\]*.

```{r}
st_geometry(mpsz)
```

### 3.2 Working with glimpse() and Date Conversion

Beside the basic geospatial feature information, I would also like to learn more about the associated attribute information in the data frame.

```{r}
glimpse(mpsz)
```

The *glimpse()* report reveals the data type of each fields. For example, `SUBZONE_NO` is in integer data type, `SUBZONE_N`, `CA_IND`, `PLN_AREA_N` and `REGION_N` fields are in character data type, and `SHAPE_1.AREA` and `SHAPE_1.LEN` fields are in double-precision values.

It should be noted that `FMEL_UPD_D` is showing as a 14-digit character rather than a date type. The 14 digits should be referring to the year, month, date, hour, minute and second in a "YYYYMMDDHHMMSS" format and will need to be converted using *ymd_hms()* within a *mutate()* function, which is part of the tidyverse package. I also dropped the time using *as_date()*.

```{r}
mpsz <- mpsz %>%
  mutate(FMEL_UPD_D = as_date(ymd_hms (FMEL_UPD_D))) #convert to date and drops the time

glimpse(mpsz) #verify that it has been converted to date
```

By executing the *glimpse()* function again, I was able to check that `FMEL_UPD_D` has indeed been converted to a date type.

### 3.3 Working with head()

To reveal complete information of a feature object, I will be using *head()*. I have indicated *n=5* to display only the first five rows.

```{r}
head(mpsz, n=5)
```

## 4. Plotting the Geospatial Data

In geospatial data science, it is not enough to just look at the feature information. Instead, we should visualise the geospatial features using *plot()* of R Graphic.

```{r}
plot(mpsz)
```

The default plot of an sf object is a multi-plot of all attributes, up to a reasonable maximum as shown above. Instead, we can choose to plot only the geometry by using the code chunk below.

```{r}
plot(st_geometry(mpsz))
```

We can also choose to plot the sf object by using a specific attribute.

```{r}
plot(mpsz["PLN_AREA_N"])
```

I will now plot the `preschool` layer ontop of the `mpsz` layer.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), 
     add = TRUE)
```

As both `preschool` and `mpsz` are in **WGS84**, the preschool points appear on the subzone map. However, the points are too dense and it is difficult to have a detailed interpretation of the exact locations within each subzone. Regardless, it provides a good overview of where preschools are generally not located, namely the industrial, protected nature areas and offshore islands.

## 5. Working with Projection 
