---
title: "01b: Thematic Mapping and GeoVisualisation with R"
author: "Chia Ka Ying"
date: "2026-08-26"
date-modified: "2026-08-27"
---

Thematic mapping involves the use of map symbols to visualise properties of geographic features that are not naturally visible, such as population, temperature, crime rate and property prices.

Geovisualisation, on the other hand, is the process of using visual representations and cartographic techniques to explore, analyse and communicate geospatial data. It combines elements of cartography, computer science and information visualisation to enhance spatial understanding and knowledge discovery.

In this chapter, I will plot functional and truthful choropleth maps by using an R package called **tmap** package.

## 1. Importing Packages and Data

### 1.1 Importing packages

Aside from the **tmap** package, I will also be importing **sf**, **tidyverse** and **rvest**.

```{r}
pacman::p_load(sf, tmap, tidyverse, rvest)
```

### 1.2 Importing data into R

I will be using the below two data sets to create the choropleth map:

- Master Plan 2019 Subzone Boundary (No Sea). This shows the geographical boundary of Singapore at the planning subzone level and is based on URA Master Plan 2019. The format is in GEOJSON and I will be converting it to KML format.

- Singapore residents by planning area / subzone, age group, sex and type of dwelling, June 2024 in csv format. This is an aspatial data file downloaded from Department of Statistics, Singapore. Although it does not contain any coordinates values, its `PA` and `SZ` fields can be used as unique identifiers to georeference to the planning subzone data set.

### 1.3 Importing geospatial data into R

Using the *st_read()* of sf package, I will be importing the polygon feature layer in **GeoJSON** format file `MP14_SUBZONE_WEB_PL` into R. will also be converting it from a **GeoJSON** format to **KML** format using the *st_write()* function. To ensure that any existing file is not overwritten, I have included a *delete_dsn = FALSE* argument and *#\| eval: false* chunk option to display the code in the final rendered code but prevents RStudio from executing it during the rendering process to prevent an error.

```{r}
mpsz <- st_read("data/geospatial/MP14_SUBZONE_NO_SEA_PL.geojson")
```

```{r}
#| eval: false #prevent error when rendering

dir.create("data/geospatial/mpsz", recursive = TRUE, showWarnings = FALSE) #create folder

st_write(mpsz, "data/geospatial/mpsz/MP14_SUBZONE_NO_SEA_PL.kml", 
         driver = "KML", delete_dsn = FALSE) #write files in KML format and prevent overwrite
```

Tidying is required for KML files because their attribute data is embedded as an HTML table within a single description field, rather than being split into individual columns. The extraction step is needed to parse out values like REGION_N and SUBZONE_N into proper, usable columns. In my case, however, `mpsz` was imported from in GEOJSON format, which already provides `REGION_N`, `PLN_AREA_N`, `SUBZONE_N`, and `SUBZONE_C` as standalone columns, so this tidying step is unnecessary and can be skipped.

Instead, I will examine the content of mpsz. By default, it will display only the first 10 rows. To see more or less rows, you can use the `n=` argument.

```{r}
mpsz
```

### 1.4 Importing attribute data into R

I will now import the `respopagesextod2024.csv` file into RStudio and save the file into a tibble dataframe called `popdata`.

The task will be performed by using *read_csv()* function of readr package.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

### 1.5 Data preparation

Before preparing a thematic map, I will prepare a data table with values from year 2020. The data table should include the variables `PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL` and `DEPENDENCY`.

- YOUNG: age group 0 to 4 until age groyup 20 to 24,
- ECONOMY ACTIVE: age group 25-29 until age group 60-64,
- AGED: age group 65 and above,
- TOTAL: all age group, and
- DEPENDENCY: the ratio between young and aged against economy active group

I will be using the following data wrangling and transformation functions:

- pivot_wider() of tidyr package, and
- mutate(), filter(), group_by() and select() of dplyr package

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

Before joining the datasets, I will convert the `PA` and `SZ` fields to uppercase. Their values currently mix upper- and lowercase, whereas `SUBZONE_N` and `PLN_AREA_N` use uppercase letters.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, I will perform the join using *left_join()* from dplyr, using the subzone names `SUBZONE_N` and `SZ` as the common identifier to combine the geographical and attribute data.

```{r}
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                          by = c("SUBZONE_N" = "SZ"))
```

It should be noted that passing the `mpsz` simple feature data frame as the left table in *left_join()* ensures that the resulting output retains its spatial properties and remains a simple feature object.

I will now save the R object `mpsz_pop2024` as a .rds file, so that I can reload it later without having to redo all the earlier processing steps.

```{r}
write_rds(mpsz_pop2024, "data/geospatial/mpsz_pop2024.rds")
```

## 2. Choropleth Mapping Geospatial Data Using tmap

Choropleth mapping uses area patterns or graduated colors to visualise data across geographic units like countries, states or subzones.

In tmap, I can create thematic maps using two approaches:

- Quick maps with *qtm()*: Fast and simple for default visualisations.

- Custom maps with tmap elements: Offers full control over design and layering.

### 2.1 Plotting a choropleth map quickly by using qtm()

*qtm()* is the easiest way to quickly draw a choropleth map. It usually provides a good default visualisation. I will be producing a static map using the *plot* option, whereas an interactive one can be produced using the *view* option. I will also be using `DEPENDENCY` as the attribute in *fill=*.

```{r}
tmap_mode("plot")
qtm(shp = mpsz_pop2024, 
    fill = "DEPENDENCY")
```

### 2.2 Creating a choropleth map by using tmap’s elements

While *qtm()* is great for creating choropleth maps quickly, it makes fine-tuning layer aesthetics difficult. For higher-quality, cartographic-standard maps, build your visualisation using tmap's individual drawing elements instead.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by Planning Subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and population data from Department of Statistics (DOS)", 
             position = c("left", "bottom"))
```

### 2.3 Drawing a base map

The core building block of tmap starts with *tm_shape()* to define your input dataset, followed by one or more layer elements such as *tm_polygons()*, *tm_symbols()*, *tm_lines()*, *tm_raster()* or *tm_text()*.

I will be using *tm_shape()* to define the input data and *tm_polygons()* to draw the planning subzone polygons. By default, the areas of polygons is in light gray (gray85) and polygons borders are in slightly dark gray (gray25).

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

### 2.4 Drawing a choropleth map using tm_polygons()

I will be drawing a choropleth map showing the geographical distribution of a selected variable by planning subzone. The variable `DEPENDENCY` will be assigned to *tm_polygons()*.

```{r}
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY")
```

### 2.5 Drawing a choropleth map using tm_fill() and tm_border()

*tm_polygons()* is infact a wraper of *tm_fill()* and *tm_border()*. *tm_fill()* is to fill the polygons using the default colour scheme and *tm_borders()* adds the borders of the polygon features. Here is what the map would look like without and with *tm_borders()*.

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY")
```

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders()
```

We can also change the border colour, border line width and border line type accordingly.

```{r}
tm_shape(mpsz_pop2024)+
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60",
             lwd = 0.1,
             lty = "dashed")
```

### 2.6 Data classification methods of tmap

Choropleth maps group large datasets into distinct classes or ranges using data classification techniques.

The tmap package offers ten classification methods: fixed, sd, equal, pretty (the default), quantile, kmeans, hclust, bclust, fisher and jenks.

The style arguments *tm_fill()* or *tm_polygons()* can be used to define a data classification method.

#### 2.6.1 Plotting choropleth maps with built-in classification methods

I will be using the style argument of *tm_fill()* or *tm_polygons()* to define a data classification method. Here is how a quantile data classification using 5 classes is done.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Here is where I use an equal data classification method.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

For this instance, we will notice that the distribution using the quantile data classification method is more evenly distributed:

- Quantile Classification: Divides the total number of geographic features (polygons) into equal-sized groups. If you request 5 classes, it places roughly 20% of the subzones into each group. Because it assigns a fixed count of features to every class, the resulting colors are distributed evenly across the map regardless of skew in the underlying numbers.

- Equal Interval Classification: Divides the range of data values (from minimum to maximum value) into equal-sized numerical intervals. It ignores how many polygons fall into each interval. If your data contains extreme outliers or is heavily skewed toward low values, most subzones end up clumped into just one or two color categories, leaving the remaining classes nearly empty.

#### 2.6.2 DIY 1: Comparing Data Classification Methods

Here, I will compare four different data classification methods (quantile, equal, jenks and sd) using a fixed 5-class scale to see how class boundary calculations change the spatial distribution of the `DEPENDENCY` ratio.

```{r}
map_quantile <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile")

map_equal <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal")

map_jenks <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(style = "jenks", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Jenks")

map_sd <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(style = "sd", n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Standard Deviation")

# arrange in a 2x2 grid layout
tmap_arrange(map_quantile, map_equal, map_jenks, map_sd, ncol = 2)
```

The following were observed from the four maps:

- Quantile: Spreads subzones evenly across the color bins, so the map shows a nice mix of colors. The downside is it can hide extreme cases or make small, unimportant differences look bigger than they are.
- Equal Interval: Splits the value range into equal-sized chunks. Since a few subzones have unusually high dependency ratios, most other subzones get squeezed into the lowest bin. This results in the map ending up mostly with one color, with barely anything in the higher bins.
- Jenks (Natural Breaks): Groups similar values together and keeps different groups apart, so it naturally finds where the real breaks in the data are. This makes it good at showing genuine clusters, especially when the data is not evenly spread out.
- Standard Deviation: Shows which subzones sit unusually far above or below the average dependency ratio. This is useful for quickly spotting the outliers.

#### 2.6.3 DIY 2: Comparing Different Number of Classes

Using a single classification method (quantile), I will vary the number of classes (n = 2, 6, 10, 20) to observe how it affects map readability and data interpretation.

```{r}
map_c2 <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(style = "quantile", n = 2)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("2 Classes")

map_c6 <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(style = "quantile", n = 6)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("6 Classes")

map_c10 <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(style = "quantile", n = 10)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("10 Classes")

map_c20 <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY", 
              fill.scale = tm_scale_intervals(style = "quantile", n = 20)) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("20 Classes")

# Arrange in a 2x2 grid layout
tmap_arrange(map_c2, map_c6, map_c10, map_c20, ncol = 2)
```

The following were observed from the four maps:

- 2 Classes: Splits everything into just "high" or "low," which is easy to read at a glance but loses the details.
- 6 Classes: Seems to strike a balance between visual clarity and details.
- 10–20 Classes: The shades are become too subtle in differences, making it difficult to visually differentiate.

#### 2.6.4 Plotting choropleth map with custom break

By default, tmap calculates category breaks automatically. To set custom intervals instead, we will need to indicate values to the breaks argument in *tm_scale_intervals()*. Before choosing custom breakpoints, it is best practice to inspect the data’s summary statistics.

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

Based on the result, I will set break points at at 0.60, 0.70, 0.80, and 0.90, and include a minimum and maximum of 0 and 100. The breaks vector will be c(0, 0.60, 0.70, 0.80, 0.90, 1.00).

```{r}
tm_shape(mpsz_pop2024)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5)
```

### 2.7 Colour scheme

tmap also supports colour ramps as defined by users or based on a set of predefined colour ramps from the RColorBrewer package.

#### 2.7.1 Using ColorBrewer palette

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

We will need to add a "-" prefix to reverse the shading of the map based on the `DEPENDENCY` attribute.

```{r}

tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

#### 2.7.2 Cartographic Furniture

tmap also can be used to draw other map furnitures, such as compass, scale bar, grid lines and data sources using *tm_compass()*, *tm_scale_bar()*, *tm_grid()* and *tm_credit()*.

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

### 2.8 Map Layout

Map layout is the coming together of map elements to form a cohesive map. This can be customised using the *tm_layout()* function.

#### 2.8.1 Map Legend

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_pos_auto_in() +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

#### 2.8.2 Map Style

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) + 
  tm_borders(fill_alpha = 0.5) + 
  tmap_style("natural")
```

## 3. Drawing Small Multiple Choropleth Maps

Maps can also be stacked vertically, arranged side-by-side or in grids, as demonstrated in 2.6.2 and 2.6.3. This allows the visualisation of relationships change based on a variable, such as time.

### 3.1 By assigning multiple values to at least one of the aesthetic arguments

By assigning two variables to the visual variable.

```{r}
tm_shape(mpsz_pop2024) + 
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = 
      tm_legend(position = tm_pos_in(
        "right", "bottom")),
    fill.scale = tm_scale_intervals(
      style = "equal", 
      n = 5,
      values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

### 3.2 By arranging multiples choropleth maps in a grid layout

```{r}
youngmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                  item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")
                
agedmap <- tm_shape(mpsz_pop2024)+ 
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(
                position = tm_pos_in(
                  "right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
              style = "quantile", 
              values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### 3.3 By defining a group-by variable in tm_facets()

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) + 
  tm_facets(by = "REGION_N",
            nrow = 2, 
            ncols = 3,
            free.coords=TRUE, 
            drop.units=TRUE) +
  tm_layout(legend.show = TRUE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(fill_alpha = 0.5)
```

## 4. Mappping Spatial Object Meeting a Selection Criterion

*filter()* can be used to select a certain geographical area and to plot a choropleth map on the selected region. I will plot this based on `CENTRAL REGION`.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

## 5. Complementing Thematic Map with Statistical Chart

Maps and statistical charts can complement each other to represent different aspects of the same data, thus enhancing a reader's understanding.

tmap can be used to include statistical chart into the map visualisation by using the *fill.chart* argument to include box plot.

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile", 
                values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

I will be improving the visual representation by highlighting and labelling the outliers on the choropleth map.

```{r}
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected$DEPENDENCY)

outlier_vals <- stats$out

outlier_sf <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected) +
  tm_polygons(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile", 
            values = "brewer.blues"),
          fill.legend = tm_legend(),
          fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
tm_shape(outlier_sf) +
  tm_borders(col = "darkred", lwd = 3) +
  tm_text("SUBZONE_N", col = "red", size = 1) +
  tm_layout(asp = 0.8)
```

## 6. Creating Interactive Map

Interactive maps are useful in letting users explore the data based on their requirements and to gain a better understanding of the information. *tmap_mode()* allows you to switch easily between static and interactive maps.

I will modify the code from section 4 to build an interactive map.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2)
```

Without limiting the *set_zoom_limits* argument, users might have difficulty navigating the map.

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")

tm_shape(region_selected, 
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name" = "SUBZONE_N", 
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12,14))
```

## 7. Reference

Kam, T. S. Thematic Mapping and GeoVisualisation with R. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap02#creating-a-choropleth-map-by-using-tmaps-elements
