---
title: "01a: Geospatial Data Science with R"
---

Geospatial data science is a process of importing, wrangling, integrating, and processing geographically referenced data sets. In this hands-on exercise, I will perform geospatial data science tasks in R by using sf package.

## 1. Getting Started: Loading R Packages

Using the *p_load* function of pacman package, I will load sf and tidyverse R packages into the R environment.

```{r}
pacman::p_load(sf, tidyverse)
```

## 2. Importing Geospatial Data

Using the *st_read()* of sf package, I will be loading the following three geospatial data and converting them to the desired formats:

- `MP14_SUBZONE_NO_SEA_PL`, a polygon feature layer in GeoJSON format, which will be converted to an ESRI shapefile format;

- `PreSchoolsLocation`, point feature layer in GeoJSON format, which will be converted to a KML file format; and

- `CyclingPath`, a line feature layer in ESRI shapefile format.

### 2.1 Importing polygon feature data and coverting to shapefile format (MP14_SUBZONE_NO_SEA_PL)

Using the *st_read()* function of sf package, I will be importing the polygon feature layer in **GeoJSON** format file `MP14_SUBZONE_WEB_PL` into R. I will also be converting it from a **GeoJSON** format to an **ESRI shapefile** format using the *st_write()* function. To ensure that any existing file is not overwritten, I have included a *delete_dsn = FALSE* argument and *#\| eval: false* chunk option to display the code in the final rendered code but prevents RStudio from executing it during the rendering process to prevent an error.

```{r}
mpsz <- st_read("data/geospatial/MP14_SUBZONE_NO_SEA_PL.geojson")
```

```{r}
#| eval: false #prevent error when rendering

dir.create("data/geospatial/mpsz", recursive = TRUE, showWarnings = FALSE) #create folder

st_write(mpsz, "data/geospatial/mpsz/MP14_SUBZONE_NO_SEA_PL.shp", 
         driver = "ESRI Shapefile", delete_dsn = FALSE) #write files in ESRI format and prevent overwrite
```

The message above reveals that the geospatial objects are multipolygon features. There are a total of 323 multipolygon features and 13 fields in `mpsz` simple feature data frame. `mpsz` is in **WGS 84**, which is a geodetic system and thusgeographic (3D/spherical) rather than projected (2D/flat). The bounding box provides the x extend and y extend of the data.

### 2.2 Importing point feature layer and coverting to KML format (PreSchoolsLocation)

Similary, using the *st_read()* function of sf package, I will be importing the point feature layer in **GeoJSON** format file `PreSchoolsLocation` into R. I will also be converting it from a **GeoJSON** format to **KML** format using the *st_write()* function. To ensure that any existing file is not overwritten, I have included a *delete_dsn = FALSE* argument and *#\| eval: false* chunk option to display the code in the final rendered code but prevents RStudio from executing it during the rendering process to prevent an error.

```{r}
preschool <- st_read("data/geospatial/PreSchoolsLocation.geojson")
```

```{r}
#| eval: false #prevent error when rendering

dir.create("data/geospatial/preschool", recursive = TRUE, showWarnings = FALSE) #create folder

st_write(preschool, "data/geospatial/preschool/preschool.kml", 
         driver = "KML", delete_dsn = FALSE) #write files in KML format and prevent overwrite
```

The message above reveals that `preschool` is a point feature data frame. There are a total of 2290 features and 2 fields. `preschool` is in **WGS 84**, which is the same as `mpsz`.

### 2.3 Importing polyline feature data in shapefile format (CyclingPath)

The code chunk below uses *st_read()* function of sf package to import CyclingPath shapefile into R as a line feature data frame.

```{r}
cyclingpath = st_read(dsn = "data/geospatial/CyclingPath", 
                         layer = "CyclingPath")
```

There are a total of 4830 features and 7 fields in cyclingpath sf data frame. The geospatial entities are capture in multilinestring object. Unlike the earlier two data sets, this data set is in **SVY21** projected coordinates system.

## 3. Checking the Content of a Simple Feature Data Frame

Before performing data wrangling, I will be learning more about the `mpsz` data set using *st_geometry(),* *glimpse()* and *head().*

### 3.1 Working with st_geometry

To retrieve the geometry list-column, I will be using the below code chunk, which is the more general way, rather than *mpsz\$geom* or *mpsz\[\[1\]\]*.

```{r}
st_geometry(mpsz)
```

### 3.2 Working with glimpse() and Date Conversion

Beside the basic geospatial feature information, I would also like to learn more about the associated attribute information in the data frame.

```{r}
glimpse(mpsz)
```

The *glimpse()* report reveals the data type of each fields. For example, `SUBZONE_NO` is in integer data type, `SUBZONE_N`, `CA_IND`, `PLN_AREA_N` and `REGION_N` fields are in character data type, and `SHAPE_1.AREA` and `SHAPE_1.LEN` fields are in double-precision values.

It should be noted that `FMEL_UPD_D` is showing as a 14-digit character rather than a date type. The 14 digits should be referring to the year, month, date, hour, minute and second in a "YYYYMMDDHHMMSS" format and will need to be converted using *ymd_hms()* within a *mutate()* function, which is part of the tidyverse package. I also dropped the time using *as_date()*.

```{r}
mpsz <- mpsz %>%
  mutate(FMEL_UPD_D = as_date(ymd_hms (FMEL_UPD_D))) #convert to date and drops the time

glimpse(mpsz) #verify that it has been converted to date
```

By executing the *glimpse()* function again, I was able to check that `FMEL_UPD_D` has indeed been converted to a date type.

### 3.3 Working with head()

To reveal complete information of a feature object, I will be using *head()*. I have indicated *n=5* to display only the first five rows.

```{r}
head(mpsz, n=5)
```

## 4. Plotting the Geospatial Data

Using *plot()*, you can visualise a data frame by generating coloured maps based on different attributes

```{r}
plot(mpsz)
```

The default plot of an sf object is a multi-plot of all attributes, up to a reasonable maximum as shown above. This includes visualisation of non-informative features, such as `OBJECTID_1`.

Instead, you can plot only the geometry by using *st_geometry()* where only the shape and not the attributes are mapped out.

```{r}
plot(st_geometry(mpsz))
```

We can also plot using a specific attribute, such as `PLN_AREA_N`.

```{r}
plot(mpsz["PLN_AREA_N"])
```

When you have more than one layer, you can plot them ontop of each other to visualise more information. I will now plot `preschool` layer ontop of the `mpsz` layer.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), 
     add = TRUE)
```

As both `preschool` and `mpsz` are in **WGS84**, the preschool points appear on the subzone map. However, the points are too dense and it is difficult to have a detailed interpretation of the exact locations within each subzone. Regardless, it provides a good overview of where preschools are generally not located, namely the industrial areas, nature reserves and offshore islands.

Let's now try plotting the `cyclingpath` layer ontop of the `mpsz` layer, where the former is in **SVY21** while the latter is in **WGS84**.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(cyclingpath),
     add = TRUE, col = "Blue")
```

Cycling path lines do not appear on the subzone layer due to the different coordinate system used. Hence, projection transformation will need to be performed in order to geoprocess these two sets of data.

## 5. Working with Projection

### 5.1 Transforming projection of mpsz and preschool from WGS84 to SVY21

When analysis of geospatial data involves the use of distance and/or area measurement, we will need to transform data from geographic coordinate system to projected coordinate system. The transformation uses the *st_transform()* package, where the reprojection to another coordinate system is done mathematically.

SVY21 is Singapore's official national horizontal geodetic datum and plane coordinate system, and is based on the WGS84 (World Geodetic System 1984) to flatten the curved surface of Earth into metric measurements. We will be referencing an EPSG code to identify specific coordinate reference system (CRS), which in the case of SVY21 is 3414.

```{r}
mpsz <- st_transform(mpsz, 
                              crs = 3414)

st_geometry(mpsz)
```

```{r}
preschool <- st_transform(preschool, 
                              crs = 3414)

st_geometry(preschool)
```

I will also transform `cyclingpath` to the same EPSG: 3414.

```{r}
cyclingpath <- st_transform(cyclingpath, crs = 3414)
```

Projection transformation to SVY21 have been completed for `mpsz` and `preschool`, and they are now ready to be geoprocessed with `cyclingpath` and included for analysis involving distance and area measurements.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(cyclingpath),
     add = TRUE, col = "Blue")
```

`cyclingpath` has now successfully been layered ontop of `mpsz`, where cycling paths are shown in blue on the subzone map.

## 6. Working with Aspatial Data

### 6.1 Importing aspatial data

When working with geospatial analysis, we may also process them with aspatial data, where it is not a geospatial data but contains x- and y- coordinates in its data field. I will be using `listings` obtained from Inside Airbnb, which contains listings of Airbnb in Singapore.

Since `listings` is in a csv file format, I will use *read_csv()* from the readr package to import the data set as a tibble data frame called `airbnblistings`. I will also use Base R *list()* to preview the first 10 rows, which is similar to the use of *glimpse()*.

```{r}
airbnblistings <- read_csv("data/aspatial/listings.csv")

list(airbnblistings)
```

From the output, `airbnblistings`contains 3247 rows and 19 columns. We will be focusing on the fields `latitude` and `longtitude`, which are in decimal degree format. Based on a quick Google search of the numbers, they fall within the range of WGS 84 coordinate system for Singapore, which uses a latitude range from -90° to +90° and a longitude range from -180° to +180°.

### 6.2 Creating a simple data frame from an aspatial data frame

I will now convert `airbnblistings` into a simple data frame using *st_as_sf()* of sf package. The argument is detailed as follows:

- coords argument requires the column name of the x-coordinates (longitude) first, followed by the y-coordinates (latitude).
- crs argument requires the coordinates system in EPSG format. EPSG: 4326 is WGS84 Geographic Coordinate System and EPSG: 3414 is Singapore's SVY21 Projected Coordinate System. Other countries' EPSG code by found on epsg.io.
- %\>% is used to nest *st_transform()* to transform the newly created simple feature data frame into the SVY21 Projected Coordinate System.

```{r}
airbnblistings_sf <- st_as_sf(airbnblistings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

I will now examine the content of the newly created simple feature data frame `airbnblistings_sf` using *glimpse()*.

```{r}
glimpse(airbnblistings_sf)
```

With the transformation, notice that a new column `geometry` using projected coordinate reference system is created. You will also notice that the fields `latitude` and `longtitude` have been dropped.

I will now plot `airbnblistings_sf` ontop of `mpsz` to visualise Airbnb listings in Singapore subzones.

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(airbnblistings_sf), add = TRUE)
```

## 7. Geoprocessing with sf package

Besides functions to handle geospatial data, such as importing, assigning projections and transformations which were shown earlier, sf package offers a wide range of geoprocessing functions. This is also known as Geographic Information System (GIS) analysis.

### 7.1 Use Case 1: land acquisition analysis

**Scenario** The authority is planning to upgrade the exiting cycling path. To do so, they need to acquire 5 meters of reserved land on both sides of the existing cycling path. You are tasked to determine the extend of the land that needs to be acquired and the total area.

**Solution** First, I will use *st_buffer* to compute the 5-meter buffer around the cycling paths. The function is used to expand or shrink vector geometries outward by a specified distance, where the width is defined by the *dist=* referencing the metrics of SVY21. *nQuadSegs* controls the smoothness of and must be an integer more than or equals to 1. Although there is no upper limit, high values may produce negligible visual improvement.

```{r}
buffer_cycling <- st_buffer(
  cyclingpath, dist=5, nQuadSegs = 30)
```

I will now calculate the area of the buffer using tidyverse.

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

Lastly, I will use *sum()* to derive the total land needed

```{r}
sum(buffer_cycling$AREA)
```

The buffer can also be plotted based on a selected planning subzone for the purpose of land acqusition planning, such as Tampines West .

First, I will use *filter()* to extract the polygon feature of Tampines West.

```{r}
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST")
```

Next, I will use *st_intersection()* to clip the cycling buffers within Tampines West planning subzone.

```{r}
buffer_cycling_selected <- st_intersection(
  buffer_cycling, mpsz_selected)
```

Lastly, I will use *plot()* to create the plot as shown below.

```{r}
plot(st_geometry(buffer_cycling_selected))
```

### 7.2 Use case 2: To determine the number of pre-schools by planning subzone

**Scenario** The authority requires a count of pre-schools for each planning subzone to support forward planning. Using R and the sf package, perform the necessary geoprocessing to compute these counts and present the results clearly.

**Solution** First, to identify pre-schools located inside each Planning Subzone by using *st_intersects()*. Next, *length()* is used to calculate numbers of pre-schools that fall inside each planning subzone.

```{r}
mpsz$`PreSch Count`<- lengths(st_intersects(mpsz, preschool))
```

I will now check the summary statistics of the newly derived `PreSch Count` field by using *summary()*.

```{r}
summary(mpsz$`PreSch Count`)
```

To list the planning subzone with the most number of pre-school, the *top_n()* of dplyr package is used.

```{r}
top_n(mpsz, 1, `PreSch Count`)
```

Another geoprocessing function of sf package called *st_area()* is used to derive the area of each planning subzone.

```{r}
mpsz$Area <- mpsz %>%
  st_area()
```

Next, I will use *mutate()* to compute the density.

```{r}
mpsz <- mpsz %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

I will now visualise the derived variables by using exploratory data analysis methods from ggplot2.

First, I will plot a histogram to reveal the distribution of `PreSch Density` using *hist()* of R Graphics.

```{r}
hist(mpsz$`PreSch Density`)
```

Although the code chunk is easy to use, the resulting visualisation is not of publication quality. Improvements could be made by including a title and subtitle, having proper x and y labels, defining the bins for better division of data and including colour.

```{r}
ggplot(data=mpsz, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="yellow") +
  labs(title = "Are pre-school evenly distributed in Singapore?",
       subtitle= "There are many planning subzones with a single pre-school. On the other hand, \nthere are seven planning subzones with at least 30 or more pre-schools.",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

I will now use ggplot2 functions to plot a scatterplot showing the relationship between `Pre-school Density` and `Pre-school Count`.

```{r}
ggplot(data=mpsz, 
       aes(y = `PreSch Count`, 
           x= as.numeric(`PreSch Density`)))+
  geom_point(color="black", 
             fill="black") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
      x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```

## 8. Reference

Kam, T. S. Geospatial Data Science with R. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap01.html#importing-the-aspatial-data