---
title: "03: Spatio-Temporal Point Patterns Analysis"
author: "Chia Ka Ying"
date: "2026-09-05"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

Spatio-temporal point patterns capture the interplay between location and timing in real-world occurrences, from environmental disasters to ecological changes. By examining events through both spatial and temporal lenses, we can better understand how localised disturbances form, spread and interact over time. Using various R packages, this assignment explores forest fire occurrences in Kepulauan Bangka Belitung, Indonesia from 1 January 2023 to 31 December 2023 to run a set of spatio-temporal point pattern analyses.

The research questions for this assignments are as follows:

- Are the locations of forest fire in Kepulauan Bangka Belitung spatial and spatio-temporally independent?
- If the answer is no, where and when the observed forest fire locations tend to cluster?

The following data sets will be used:
- `forestfires`, a csv file provides locations of forest fire detected from the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor data. The data are downloaded from Fire Information for Resource Management System. The data has forest fires from the whole Indonesia in 2023. For the purpose of this exercise, only forest fires within Kepulauan Bangka Belitung will be used.
- `Kepulauan_Bangka_Belitung`, an ESRI shapefile showing the sub-district boundary of Kepulauan Bangka Belitung. The data set was downloaded from the Indonesia Geospatial portal. The website has a shapefile for the whole Indonesia and a geological shapefile of the Bangka Belitung Islands Province. I downloaded the geological shapefile as it is the required location for the exercise.

## 1. Importing and Wrangling Geospatial Data Sets

I will first install and load the R packages.

```{r}
pacman::p_load(sf, spatstat, sparr, 
               tmap, stpp, tidyverse)
```

### 1.1 Checking Data Set

```{r}
kbb <- st_read(dsn = "data/Bangka_Belitung",
               layer = "Geology Bangka Belitung") %>%
  st_transform(crs = 32748)
```
::: callout-note
## Note

Looking at the details of the shapefile, there are a number of differences with that in the exercise guide:

- Has only 33 features and 12 fields, compared to 297 features and 26 fields.
- Geometry type is multipolygon instead of polygon.
- Dimension is XY instead of XYZ
- Bounding box is "xmin: 104.9592 ymin: -3.961192 xmax: 109.1553 ymax: -1.477503" instead of "xmin: 105.1085 ymin: -3.116593 xmax: 106.8488 ymax: -1.501603".

As the exercise uses *st_union* to dissolve the features into a single polygon and drops attribute fields, and the subsequent `kbb_sf` and `kbb_owin` is used as a boundary to show the fire points, I might be able to continue using this geological shapefile. 

I will first check the attribute table to check the names of the 12 fields and plot it to see what it looks like with the areas labelled based on the field "NM_LEMBAR", which was identified through trial and error.

:::

```{r}
names(kbb)
```

```{r}
tm_shape(kbb) + tm_polygons(col = "NM_LEMBAR")
```

::: callout-note
## Note

I managed to find out more about the locations based on a search:

JAMBI:	Jambi province — southern Sumatra mainland
LAHAT:	Lahat, South Sumatra — mainland
TULUNG SELAPAN:	Ogan Komering Ilir, South Sumatra — mainland
PANGKALPINANG:	Capital city, on Bangka Island
SUNGAILIAT:	Regency seat of Bangka Regency, on Bangka Island
TANJUNG PANDAN:	Regency seat of Belitung Regency, on Belitung Island

This means that I am likely to require only PANGKALPINANG and SUNGAILIAT based on the plotted map in the exercise guide.

I will plot the fire incidents to confirm the locations to drop.

:::

```{r}
fire_sf <- read_csv("data/modis_2023_Indonesia.csv") %>%
  st_as_sf(coords = c("longitude","latitude"), crs = 4326)

tm_shape(kbb %>% filter(NM_LEMBAR %in% c("PANGKALPINANG","SUNGAILIAT"))) +
  tm_polygons(col = "NM_LEMBAR") +
tm_shape(fire_sf) +
  tm_dots()
```

### 1.2 Importing Study Area and Converting OWIN

I will now create the `kbb_sf` shapefile and `kbb_owin` owin object for map plotting. I also included the *st_make_valid* function as there was error with invalid polygon due to duplicated vertex, which could be due to it being a geological layer.

```{r}
kbb_sf <- st_read(dsn = "data/Bangka_Belitung",
               layer = "Geology Bangka Belitung") %>%
  filter(NM_LEMBAR %in% c("PANGKALPINANG", "SUNGAILIAT")) %>%
  st_make_valid() %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)

kbb_owin <- as.owin(kbb_sf)
kbb_owin
class(kbb_owin)
```

## 2. Importing and Preparing Forest Fire Data

::: callout-note
## Note

As the forest fire data set contains fire incidents for the whole Indonesia in 2023, I will use the *st_filter* function to include only the fire incidents that spatially intersects with `kbb_sf`, which is our required location.

:::

```{r}
fire_sf <- read_csv("data/modis_2023_Indonesia.csv") %>%
  st_as_sf(coords = c("longitude","latitude"), crs = 4326) %>%
  st_transform(crs = 32748) %>%
  st_filter(kbb_sf)
```

Because ppp object only accept numerical or character as mark. The code chunk below is used to convert data type of acq_date to numeric.

```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

## 3. Visualising the Fire Points

### 3.1 Overall plot

```{r}
tm_shape(kbb_sf) +
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

### 3.2 Visualising geographic distribution of forest fires by month

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.1) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

## 4. Computing STKDE by Month


## XX References

Kam, T. S. Spatio-Temporal Point Patterns Analysis. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap06

Peter Hall (1990) “Using the Bootstrap to Estimate Mean Squared Error and Select Smoothing Parameter in Nonparametric Problems”, JOURNAL OF MULTIVARIATE ANALYSIS, Vol. 32, pp. 177-203.