---
title: "02a: First Order Spatial Points Analysis Methods"
author: "Chia Ka Ying"
date: "2026-08-29"
date-modified: "2026-09-01"
---

Spatial Point Pattern Analysis (SPPA) is a framework for examining how locations and events are distributed across a geographic area. In practice, these points represent real-world occurrences—such as traffic accidents, crime incidents or disease outbreaks, as well as essential community infrastructure like healthcare facilities, childcare centers and local businesses.

First-Order Spatial Point Pattern Analysis (1st-SPPA) focuses on point intensity and overall density across space. Rather than looking at how individual points interact with one another, first-order analysis examines how point concentration varies across different parts of a region to identifying trends or patterns in point density.

1st-SPPA helps address three core questions:

- Concentration: Where are events or services most heavily concentrated?
- Variation: Is the distribution even across the area, or does it vary significantly by location?
- Dispersion: How spread out is the overall pattern across the study region?

By mapping these spatial patterns, 1st-SPPA provides an intuitive starting point for understanding community needs, evaluating accessibility and guiding resource allocation.

In this assignment, I will be using 1st-SPAA to answer two questions:

- Are the childcare centres in Singapore randomly distributed throughout the country?
- If the answer is no, then where are the locations with higher concentration of childcare centres?

I will be using the following two data sets for the analysis:

- Child Care Services data from data.gov.sg, a point feature data providing both location and attribute information of childcare centres.
- Master Plan 2019 Subzone Boundary (No Sea), a polygon feature data providing information of URA's 2019 Master Plan's planning subzone boundary data.

::: callout-note
## Note

As both data sets are available only in GEOJSON format on data.gov.sg, I will be using this format instead of KML.
:::

## 1. Importing and Wrangling Geospatial Data Sets

I will first install the relevant R packages.

```{r}
pacman::p_load(sf, terra, spatstat, 
               tmap, rvest, tidyverse)
```

I will next import the `MasterPlan2019SubzoneBoundaryNoSea` GEOJSON file as `mpsz_sf`and transform the EPSG to 3414.

::: callout-note
## Note

I did not build a function called *extract_kml_field* for extracting *REGION_N*, *PLN_AREA_N*, *SUBZONE_N* and *SUBZONE_C* as guided in the hands-on exercise, as this workflow is relevant only if the attribute fields are embedded in a single *Description* column.

Instead, upon inspecting the GeoJSON file, the four fields already exist as individual columns. I instead read the file directly using *st_read()* and kept the filtering to create `mpsz_cl`.
:::

```{r}
mpsz_sf <- st_read("data/MasterPlan2019SubzoneBoundaryNoSea.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414) %>%
  select(REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C)
```

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS") %>%
  filter(!st_is_empty(.)) %>%
  st_make_valid()

write_rds(mpsz_cl, "data/mpsz_cl.rds")
```

I will now import the Childcare Service GeoJSON file as an sf data frame called childcare_sf and transform the EPSG to 3414.

```{r}
childcare_sf <- st_read("data/ChildCareServices.geojson")%>%
  st_transform(crs = 3414)
```

## 2. Mapping the Geospatial Data Sets

After checking the reference system of the two geospatial data sets, I will plot them to observe their spatial patterns and ensure that the CRS matches, considering that the EPSG were both transformed earlier.

```{r}
tm_shape(mpsz_cl) +
  tm_polygons() +
tm_shape(childcare_sf) +
  tm_dots()
```

The map can also be made interactive to allow users to zoom in and out to better observe the where the Childcare Centres are plotted.

```{r}
tmap_mode('view')
tm_shape(childcare_sf)+
  tm_dots()
```

::: callout-note
## Note

As each interactive mode consumes a connection, I will switch back to plot mode to avoid displaying an excessive number of interactive maps.
:::

```{r}
tmap_mode('plot')
```

## 3. Geospatial Data Wrangling

This section demonstrates how to transform sf point and polygon data frames using *as.ppp()* and *as.owin()*, preparing them for 1st- and 2nd-order spatial point pattern methods such as Kernel Density Estimation and nearest neighbor tests.

::: callout-note
## Note

Before performing spatial point pattern analysis in R, standard simple feature (sf) objects must be converted into data structures recognised by the spatstat package. Specifically, spatstat requires point events to be formatted as ppp (Planar Point Pattern) objects and the surrounding study region to be formatted as an owin (Observation Window) object.
:::

### 3.1 Converting sf data frames to ppp class

spatstat requires the point event data in ppp object form. I will now be converting `childcare_sf` to a ppp format `childcare_ppp`.

```{r}
childcare_ppp <- as.ppp(childcare_sf)
```

I will next use *class()* of Base R to verify the object class of `childcare_ppp`.

```{r}
class(childcare_ppp)
```

The output shows that it is in ppp class. We will now review the summary statistics of the newly converted ppp object.

```{r}
summary(childcare_ppp)
```

### 3.2 Creating owin object

It is a good practice to confine the analysis within a geographical area when analysing spatial point patterns, e.g., within Singapore's boundary. In spatstat, an object called owin is specially designed to represent this polygonal region.

I will be using *as.own()* to convert `mpsz_sf` into an owin object of spatstat, and checking the class.

```{r}
sg_owin <- as.owin(mpsz_cl)
```

```{r}
class(sg_owin)
```

The output shows that the class is indeed owin. I will now display the `sg_owin` object using the *plot()* function.

```{r}
plot(sg_owin)
```

### 3.3 Combining point events object and owin object

For the last step of geospatial data wrangling, I will extract childcare events located in Singapore. I will also check the output of the point and polygon feature combined in one ppp object class.

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

```{r}
childcareSG_ppp
```

::: callout-note
## Note

The "Mark variables" is different from the hands-on exercise guide, which uses a KML that exposes only the generic *Name* and *Description* fields. This does not affect subsequent analysis, as functions like kernel density estimation and CSR tests use only point coordinates and the observation window, not the marks. The "window" and "enclosing rectangle" values confirm that the underlying spatial data is identical.
:::

## 4. Clark-Evan Test for Nearest Neighbour Analysis

This section introduces the Clark-Evans test of aggregation to evaluate whether childcare services in Singapore follow a specific spatial pattern or are distributed purely by chance. As a key method in Nearest Neighbor Analysis (NNA), the Clark-Evans test measures the mean distance between each point feature and its nearest neighbour, summarising the pattern through the Clark-Evans aggregation index (R).

To test for spatial randomness, we evaluate the following two hypotheses at a 95% confidence level:

- Ho = The distribution of childcare services are randomly distributed.

- H1 = The distribution of childcare services are not randomly distributed.

By executing *clarkevans.test()* from the spatstat.explore package, we can formally quantify the spatial structure of these services to determine if they tend to group together in specific areas or spread out evenly across the country.

### 4.1 Perform the Clark-Evans test without CSR

The *clarkevans.test()* of spatstat.explore package can be performed with or without Complete Spatial Randomness (CSR). I will now do so without CSR.

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"))
```

::: {.callout-tip style="color: green;"}
## Statistical Conclusion

At a 95% confidence level, the Clark-Evans test gives a result of Clark-Evans aggregation index of R = 0.53532 and a p-value \<2.2e-16. Because the p-value is well below the significance threshold of 0.05, we reject the null hypothesis.

The value of R indicates the underlying spatial structure of the point data:

- R = 1: The spatial distribution is completely random.
- R \< 1: The spatial distribution exhibits clustering (points are closer to each other than expected by chance).
- R = 0: Indicates complete clustering (all points at the exact same location).
- R \> 1: The spatial distribution exhibits dispersion or regularity (points are spread out uniformly).

Thus, with R \< 1, we can conclude that the spatial distribution of childcare services across Singapore exhibits statistically significant spatial clustering rather than a random distribution.
:::

::: {.callout-tip style="color: blue"}
## Business Communication

The spatial analysis confirms that childcare centers in Singapore are heavily concentrated in specific geographic clusters rather than being evenly or randomly distributed across the country.

- **Targeted Expansion & Resource Allocation:** Areas with low density or longer distances between facilities may indicate underserved communities. Educational authorities and urban planners can use these findings to identify service gaps and prioritise new childcare center in emerging residential areas.

- **Competitive Analysis for Operators:** Private providers can leverage these spatial clustering patterns to identify market saturation in high-density areas and explore untapped regions where parents' demand exceeds current capacity.

However, it should be noted that childcare centers may be absent or sparse in industrial or nature reserves, as there is little or no demand in these areas.
:::

### 4.2 Perform the Clark-Evans test with CSR

I will now perform the test with CSR, using the argument method = “MonteCarlo”. Instead of assuming a standard distribution, this approach calculates the p-value by generating a specified number of simulated realisations from nsim, i.e., 99, of Complete Spatial Randomness based on the exact count of point events in our dataset. The observed Clark-Evans aggregation index R is then compared against the distribution of R values derived from these simulated CSR patterns to test our hypotheses.

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                method="MonteCarlo",
                nsim=99)
```

::: {.callout-tip style="color: green;"}
## Statistical Conclusion

Both the section 4.1 test without CSR and the Monte Carlo simulation evaluate whether childcare services in Singapore follow a Complete Spatial Randomness (CSR) pattern.

Both tests resulted in the same index of R = 0.53532. Because R \< 1, it signals spatial clustering.

Without CSR, the p-value is \<2.2e-16. With CSR, the p-value is 0.01. This is the lowest possible p-value for 99 simulations. At a 95% confidence level, both test approaches reject the null hypothesis of Complete Spatial Randomness. The test with CSR solidifies the test result, confirming that the spatial clustering of childcare centers across Singapore is statistically significant.
:::

::: {.callout-tip style="color: blue"}
## Business Communication

Similarly, spatial analysis confirms that childcare centers in Singapore are heavily concentrated in specific geographic clusters rather than being evenly or randomly distributed across the country. The test with CSR provides higher confidence for decision-makers by ensuring the observed spatial patterns hold true even under rigorous testing
:::

## 5. Kernel Density Estimation Method

Kernel Density Estimation (KDE) is an important tool in Exploratory Spatial Data Analysis (ESDA), allowing us to transform discrete spatial points—such as the locations of childcare centers, into a smooth, continuous surface of spatial intensity. Unlike traditional statistical modeling that relies on strict distributional assumptions, KDE provides a non-parametric approach to visualise real-world spatial phenomena. By mapping variations in point density across Singapore, this technique enables us to transition from asking whether clustering exists to identifying exactly where these high-density hotspots and coverage deficits are, thus allowing for deeper spatial analysis.

### 5.1 Working with automatic bandwidth selection method

I will now compute a kernel density by using the following configurations of *density()* of spatstat:

- bw.diggle() automatic bandwidth selection method. Other recommended methods are bw.CvL(), bw.scott() or bw.ppl().
- The smoothing kernel used is gaussian, which is the default. Other smoothing methods are: “epanechnikov”, “quartic” or “disc”.
- The intensity estimate is corrected for edge effect bias by using method described by Jones (1993) and Diggle (2010, equation 18.9). The default is FALSE.

```{r}
kde_SG_diggle <- density(
  childcareSG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 
```

I will now use the *plot()* function of Base R to display the kernel density derived.

```{r}
plot(kde_SG_diggle)
```

The function *summary()* of Base R is used to print the summary report.

```{r}
summary(kde_SG_diggle)
```

I will also retrieve the bandwidth used to compute the kde layer.

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

### 5.2 Rescalling KDE values

I will now use *rescale.ppp()* ito covert the unit of measurement from meter to kilometer. This addresses the issue of SVY21 unit of measurement being in meter, which is too small to comprehend in the earlier plotted map in section 5.1.

```{r}
childcareSG_ppp_km <- rescale.ppp(
  childcareSG_ppp, 1000, "km")
```

I will now re-run *density()* using the re-scaled data set and plot the output kde map.

```{r}
kde_childcareSG_km <- density(childcareSG_ppp_km,
                              sigma=bw.diggle,
                              edge=TRUE,
                              kernel="gaussian")
```

```{r}
plot(kde_childcareSG_km)
```

The map looks similar to the earlier one, but note the change in the legend.

### 5.3 Working with different automatic bandwidth methods

Besides *bw.diggle()*, there are three other spatstat functions that can be used to determine the bandwidth: *bw.CvL()*, *bw.scott()*, and *bw.ppl()*.

```{r}
bw.CvL(childcareSG_ppp_km)
```

```{r}
bw.scott(childcareSG_ppp_km)
```

```{r}
bw.ppl(childcareSG_ppp_km)
```

```{r}
bw.diggle(childcareSG_ppp_km)
```

Selecting an appropriate bandwidth is critical in KDE, as it directly determines the level of spatial smoothing applied to the point data. While automatic selection algorithms streamline this choice, different methods serve distinct analytical goals.

- Likelihood Cross-Validation (bw.ppl()): As recommended by Baddeley et al. (2016), bw.ppl() is generally preferred for spatial patterns dominated by multiple tight clusters, as it prevents over-smoothing and preserves localised density variations.

- Cross-Validation Criterion (bw.diggle()): In contrast, bw.diggle() is optimised for isolating a single, prominent cluster surrounded by random spatial noise.

I will now compare the two methods.

```{r}
kde_childcareSG.ppl <- density(childcareSG_ppp_km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```

### 5.4 Working with different kernel methods

The default kernel method used in density.ppp() is gaussian. But there are three other options, namely: Epanechnikov, Quartic and Dics.

The code chunk below will be used to compute three more kernel density estimations by using these three kernel function.

```{r}
par(mfrow=c(2,2))
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp_km, 
             sigma=0.2959712, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```

## 6. Fixed and Adaptive KDE

### 6.1 Computing KDE by using fixed bandwidth

I will compute a KDE layer by defining a bandwidth of 600 meter. The sigma value used is 0.6 as the unit of measurement of childcareSG_ppp_km object is in kilometer. This means that the 600m is 0.6km.

```{r}
kde_childcareSG_fb <- density(childcareSG_ppp_km,
                              sigma=0.6, 
                              edge=TRUE,
                              kernel="gaussian")
plot(kde_childcareSG_fb)
```

### 6.2 Computing KDE by using adaptive bandwidth

The fixed bandwidth method is very sensitive to highly skewed distribution of spatial point patterns over geographical units, e.g., urban versus rural. One way to overcome this is to use adaptive bandwidth.

I will derive the adaptive kernel density estimation by using *density.adaptive()* of spatstat.

```{r}
kde_childcareSG_ab <- adaptive.density(
  childcareSG_ppp_km, 
  method="kernel")
plot(kde_childcareSG_ab)
```

I will compare the fixed and adaptive kernel density estimation outputs.

```{r}
par(mfrow=c(1,2))
plot(kde_childcareSG_fb, main = "Fixed bandwidth")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
```

## 7. Plotting cartographic quality KDE map

### 7.1 Converting gridded output into raster

I will now convert the im (pixel image) kernal density objects into SpatRaster objects by using *rast()* of terra package and use *class()* to verify that it belongs to SpatRaster class.

```{r}
kde_childcareSG_bw_terra <- rast(kde_childcareSG_km)
```

```{r}
class(kde_childcareSG_bw_terra)
```
After confirmation, I will now look at the properties.

```{r}
kde_childcareSG_bw_terra
```

### 7.2 Assigning projection systems

As the crs property is empty, I will use *crs()* of terra to assign the CRS information on kde_childcareSG_bw_terra layer. I will review the properties again.

```{r}
crs(kde_childcareSG_bw_terra) <- "EPSG:3414"
```

```{r}
kde_childcareSG_bw_terra
```
The crs is now SVY21.

### 7.3 Plotting KDE map with tmap

I will display the raster in cartographic quality map using tmap package.

```{r}
tm_shape(kde_childcareSG_bw_terra) + 
  tm_raster(col.scale = 
              tm_scale_continuous(
                values = "viridis"),
            col.legend = tm_legend(
            title = "Density values",
            title.size = 0.7,
            text.size = 0.7,
            bg.color = "white",
            bg.alpha = 0.7,
            position = tm_pos_in(
              "right", "bottom"),
            frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1.0)
```

## 8. First Order SPPA at the Planning Subzone Level

I will conduct further analysis at the planning area level, focusing on Punggol, Tampines Chua Chu Kang and Jurong West planning areas.

### 8.1 Geospatial data wrangling

#### 8.1.1 Extracting study area

I will extract the target planning areas, i.e., Punggol, Tampines Chua Chu Kand and Jurong West.

```{r}
pg <- mpsz_cl %>%
  filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>%
  filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>%
  filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>%
  filter(PLN_AREA_N == "JURONG WEST")
```

I will now review the extracted areas by plotting.

```{r}
par(mfrow=c(2,2))
plot(st_geometry(pg), main = "Ponggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")
```

#### 8.1.2 Creating owin object

Next, I will convert these sf objects into owin objects that is required by spatstat.

```{r}
pg_owin = as.owin(pg)
tm_owin = as.owin(tm)
ck_owin = as.owin(ck)
jw_owin = as.owin(jw)
```

#### 8.1.3 Combining point events object and owin object

```{r}
childcare_pg_ppp = childcare_ppp[pg_owin]
childcare_tm_ppp = childcare_ppp[tm_owin]
childcare_ck_ppp = childcare_ppp[ck_owin]
childcare_jw_ppp = childcare_ppp[jw_owin]
```

Next, *rescale.ppp()* function is used to transform the unit of measurement from meter to kilometer.

```{r}
childcare_pg_ppp.km = rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale.ppp(childcare_jw_ppp, 1000, "km")
```

I will now plot these four areas and the locations of the childcare centres.

```{r}
par(mfrow=c(2,2))
plot(unmark(childcare_pg_ppp.km), 
  main="Punggol")
plot(unmark(childcare_tm_ppp.km), 
  main="Tampines")
plot(unmark(childcare_ck_ppp.km), 
  main="Choa Chu Kang")
plot(unmark(childcare_jw_ppp.km), 
  main="Jurong West")
```

### 8.2 Clark and Evans Test

#### 8.2.1 Choa Chu Kang planning area

I will now use *clarkevans.test()* of spatstat to performs Clark-Evans test of aggregation for childcare centres in Choa Chu Kang planning area.

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```
#### 8.2.2 Tampines planning area

I will perform a similar test for childcare centres in Tampines.

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

### 8.2 Computing KDE surfaces by planning area

I will compute the KDE of the four planning areas. bw.diggle method is used to derive the bandwidth of each area.

```{r}
par(mfrow=c(2,2))
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Jurong West")
```

## 9. Reference

Kam, T. S. First-order Spatial Point Patterns Analysis Methods. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/chap04.html