---
title: "05a: Global Measures of Spatial Autocorrelation"
author: "Chia Ka Ying"
date: "2026-09-19"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

This exercise will be computing Global Measures of Spatial Autocorrelation (GMSA) by using **spdep** package. By the end to this hands-on exercise, you will be able to:

- import geospatial data using appropriate function(s) of **sf** package,

- import csv file using appropriate function of **readr** package,

- perform relational join using appropriate join function of **dplyr** package,

- compute Global Spatial Autocorrelation (GSA) statistics by using appropriate functions of **spdep** package,

  - plot Moran scatterplot,

  - compute and plot spatial correlogram using appropriate function of **spdep** package.

- provide statistically correct interpretation of GSA statistics.

## 1. Getting Started

In spatial policy, one of the main development objective of the local government and planners is to ensure equal distribution of development in the province. In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e., GDP per capita) of Hunan Provice, People Republic of China, and to do the following:

- To apply appropriate spatial statistical methods to discover if development are evenly distributed geographically.
- If not, to find out whether there is sign of spatial clustering.
- And if yes, to find out where are these clusters.

The following data sets will be used:

Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format. Hunan_2012.csv: This csv file contains selected Hunan’s local development indicators in 2012.

I will now import the required packages to perform the study.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

## 2. Preparaing the Data

### 2.1 Importing the Data

I will now import `hunan` and `hunan2012`, which are a shapefile and csv respectively.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 2.2 Performing Relational Join

I will update the attribute table of the `hunan` SpatialPolygonsDataFrame with the attribute fields of `hunan2012` dataframe using *left_join()*. Prior to doing so, I will do a check of the column names and index for reference in the argument.

```{r}
data.frame(Index = 1:ncol(left_join(hunan, hunan2012)), 
           Name = names(left_join(hunan, hunan2012)))
```

```{r}
hunan <- left_join(hunan, hunan2012) %>%
  select(1:4, 7, 15)
```

### 2.3 Visualising Regional Development Indicator

```{r}
#| fig-width: 10
#| fig-height: 6
#| echo: false

equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fiil_alpha = 0.5) + 
  tm_layout(legend.position = c("left", "bottom"), 
            main.title = "Quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## 3. Global Measures of Spatial Autocorrelation

In this section, I will compute global spatial autocorrelation statistics and perform spatial complete randomness test for global spatial autocorrelation.

### 3.1 Computing Contiguity Spatial Weights

To compute global spatial autocorrelation statistics, a spatial weights matrix must first be constructed to define the spatial relationships and neighbourhood structure among the geographical units (i.e., counties) in the study area.

I will use the *poly2nb()* function from the spdep package to compute a contiguity-based neighbour list using shared boundary criteria. By default, the queen argument is set to TRUE, which includes all contiguous boundaries—both shared edges and single shared vertices to identify first-order spatial neighbours.

::: callout-note
## Note

Please note that 3.1 and 3.2 were also performed in Hands-On Exercise 4b if you wish to refer to how it is used Spatial Weights and Application.

:::

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```
The summary report shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbour.

## 3.2 Row-standardised weights matrix

Next, I will assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight using style=“W”. This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values. While this is the most intuitive way to summarise the neighbors’ values, the drawback is that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we will stick with the style=“W” option for simplicity’s sake but note that other more robust options are available, notably style=“B”.

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

## 4. Global Measures of Spatial Autocorrelation: Moran’s I

### 4.1 Maron’s I test

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

::: callout-note
## Note

With a p-value of < 0.01, we can reject the null hypothesis of spatial randomness. The Moran I statistic value of 0.30075 is positive, and greater than the expected value of -0.0115 under spatial randomness, showing a statistically significant positive spatial correlation. 

:::

### 4.2 Computing Monte Carlo Moran’s I

I will now use *moran.mc()* to perform permutation test, using a total of 1,000 simulations.

```{r}
set.seed(1234)
bperm= moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

::: callout-note
## Note

The results are similar even with 1,000 simulations, where the statistic is also 0.30075.

:::

### 4.3 Visualising Monte Carlo Moran’s I

I will now examine the simulated Moran’s I test statistics in greater detail. This can be achieved by plotting the distribution of the statistical values as a histogram by using the code chunk below using *hist()* and *abline()*.

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

## 5. Global Measures of Spatial Autocorrelation: Geary’s C

### 5.1 Geary’s C Test

I will perform Geary’s C test for spatial autocorrelation by using *geary.test()* of spdep.

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

### 5.2 Computing Monte Carlo Geary’s C

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

### 5.3 Visualising the Monte Carlo Geary’s C

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

## 6. Spatial Correlogram

Spatial correlograms are great to examine patterns of spatial autocorrelation in data or model residuals. They show how correlated are pairs of spatial observations when you increase the distance (lag) between them, by plotting some index of autocorrelation (Moran’s I or Geary’s c) against distance. Although correlograms are not as fundamental as variograms (a keystone concept of geostatistics), they are very useful as an exploratory and descriptive tool. For this purpose they actually provide richer information than variograms.

### 6.1 Compute Moran’s I correlogram

I will be computing a 6-lag spatial correlogram of GDPPC, using the global spatial autocorrelation in Moran's I.

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```

Plotting the output might not allow us to provide complete interpretation. This is because not all autocorrelation values are statistically significant. Hence, it is important for us to examine the full analysis report by printing out the analysis results.

```{r}
print(MI_corr)
```

### 6.2 Compute Geary’s C correlogram and plot

I will now be using *sp.correlogram()* of spdep package to compute a 6-lag spatial correlogram of GDPPC, using the global spatial autocorrelation in Geary’s C. I will also print the analysis report.

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```

```{r}
print(GC_corr)
```

