---
title: "05: Global and Local Spatial Analysis"
author: "Chia Ka Ying"
date: "2026-09-26"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

## 1. Importing R Packages

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

## 2. Preparaing the Data

### 2.1 Importing the Data

I will now import `hunan` and `hunan2012`, which are a shapefile and csv respectively.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 2.2 Performing Relational Join

I will create the dataframe `hunan_GDPPC` using `hunan` and `hunan2012` using *left_join()*. Prior to doing so, I will do a check of the column names and index for reference in the argument.

```{r}
data.frame(Index = 1:ncol(left_join(hunan, hunan2012)), 
           Name = names(left_join(hunan, hunan2012)))
```

```{r}
hunan_GDPPC <- left_join(hunan, hunan2012) %>%
  select(1:4, 7, 15, 16)
```

## 3. Global Measure of Spatial Association

#### 3.1 Computing Queen's contiguity weights

I will not compute the Queen's contiguity weights, which requires three inputs on *nb*, *style* and *allow_zero*.

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

#### 3.2 Computing Global Moran's I

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```

To ensure the computation is reproducible, we will use *set_seed()*. We will perform 100 simulations using *nsim = 99*. 

```{r}
set.seed(1234)

global_moran_perm(wm_q$GDPPC,
                      wm_q$nb,
                      wm_q$wt,
                  nsim = 99)
```

## 4. Local Measure of Spatial Association

### 4.1 Computing Local Moran's I

Note that when using sfdep package, note the p-values, `p_ii`, `p_ii_sim` and `p_folder_sim`, that are generated. Take note of the three interpretation columns `mean`, `median` and `pysal`. We should use `mean` if distribution is normal, `median` when distribution is skewed. The values in `pysal` are aligned with what is usually generated in Python. 

```{r}
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
      .before = 1) %>%
  unnest(local_moran)
```

### 4.2 Plotting LISA map

```{r}
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)
tmap_mode("plot")
tm_shape(lisa) + # this is plotted as a grey backdrop, which includes all 88 counties
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(lisa_sig) + # this is plotted to show only those that are significant
  tm_polygons(fill = "mean",
            fill.legend = tm_legend( # fill.legend places the legend box outside the map
              text.size = 1.15, # determines the legend title text size
              width = 6)) +
  tm_borders(fill_alpha = 0.4) +
  tm_layout(legend.title.size = 1.25) # determines the legend text size
```
  
  