---
title: "01: Study on Road Accidents in Bangkok Greater Metropolitan Area"
author: "Chia Ka Ying"
date: "2026-09-12"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

## 1. Introduction and Study Context

Road traffic accidents are among the most persistent and spatially patterned public-safety problems facing rapidly urbanising and urbanised regions. Unlike many other point-based public-safety events, traffic accidents are constrained to occur on the road network rather than freely across the two-dimensional landscape . This structural feature of the data has direct consequences for which analytical methods are valid.

This report investigates the spatial distribution of road traffic accidents recorded in **2022** across the **Bangkok Metropolitan Region (BMR)**, comprising Bangkok, Nonthaburi, Nakhon Pathom, Pathum Thani, Samut Prakan and Samut Sakhon. First-order and second-order spatial point pattern analysis (SPPA), with a network-constrained treatment justified by the road-bound nature of the event will be used.

## 2. Research Questions and Objectives

**Primary research questions:**

- Do road traffic accidents in the BMR occur with uniform intensity across space, or are they concentrated in identifiable areas (first-order property)?

- Do accidents exhibit evidence of spatial interaction, i.e., clustering, inhibition or approximate randomness, relative to an appropriate network-constrained null model (second-order property)?

**Objectives:**

- Prepare and integrate real-world geospatial and temporal data for point-pattern analysis.
- Investigate first-order properties of the accident point pattern across space, and, where the data allow, across time.
- Investigate second-order properties using both planar and network-constrained methods.
- Critically evaluate the assumptions, methodological choices and limitations of the methods used.
- Communicate findings through appropriate geovisualisation.
- Translate findings into public-safety planning implications.

## 3. Data and Scope of Study

## 3.1 Datasets

| \# | Dataset | Type | Role |
|-----------------|------------------|-----------------|---------------------|
| 1 | `geoBoundaries-THA-ADM1-all.shp` | Multi-polygon | Define the BMR study region (six provinces) |
| 2 | `thai_road_accident_2019_2022.csv` | Point (event) | The 2022 point events |
| 3 | `gis_osm_roads_free_1.shp` | Line | The constraining network (main roads only) |

## 3.2 Point Event Definition

A point event in this study is a single recorded road traffic accident occurring in 2022 within the BMR, represented by its reported latitude/longitude. The source data records one row per accident and includes information on the type of vehicle, accident details and casualties.

## 3.3 Data-quality handling applied to the point events:

Duplicate observations — de-duplicated on accident ID (or date+time+coordinate combination) so each physical accident is one event.

Repeated detections — not applicable to this dataset (single-sensor detections such as camera triggers are not part of this source); noted here for completeness per the report brief.

Missing coordinates — rows with NA latitude/longitude are dropped, with the count logged.

Invalid coordinates/geometries — coordinates outside plausible Thailand bounds (lat 5.5°–21°N, lon 97°–106°E) are dropped; geometries are validated with st_is_valid().

Observations outside the study area — points falling outside the dissolved six-province boundary are dropped via st_filter().

Observations outside the temporal period — only records with a parsed date falling in calendar year 2022 are retained.

## 3.4 Spatial Study Window

The spatial study window is the union of the six BMR provinces (Bangkok, Nonthaburi, Nakhon Pathom, Pathum Thani, Samut Prakan, Samut Sakhon), dissolved into a single polygon. This boundary is used both as the owin for planar spatstat analysis and as the clipping extent for the road network and traffic-infrastructure layers.

# 4. Data Preparation

```{r}
pacman::p_load(sf, terra, spatstat, spNetwork, tmap, tidyverse, dplyr, units, lubridate)
```

## 4.1 Geographical Boundary of BMR

Prior to preparing the geospatial analytical workspace, it is important to check the details of the dataset. This will aid in the extraction of the provinces in BMR into a shapefile `bmr`.

```{r}
adm1 <- st_read("data/geospatial/geoBoundaries-THA-ADM1-all/geoBoundaries-THA-ADM1.shp") %>%
  st_transform(crs = 32647)
```

```{r}
str(adm1)
```

```{r}
adm1[["shapeName"]]
```

By obtaining the values in the province field, the BMR provinces can now be filtered.

```{r}
bmr_select <- adm1 %>%
  filter(shapeName %in% c(
    "Bangkok",
    "Nonthaburi Province",
    "Pathum Thani Province",
    "Samut Prakan Province",
    "Nakhon Pathom Province",
    "Samut Sakhon Province"
  ))
```

```{r}
bmr <- bmr_select %>%
  st_union() %>%
  st_sf(name = "Bangkok Greater Metropolitan Area", geometry = .) %>%
  st_make_valid()
```

### 4.2 Road Networks in BMR

Similarly, the road networks data set will be checked in order to correctly filter the required data. For the purpose of this analysis, we will use only main roads.

```{r}
roads_raw <- st_read("data/geospatial/thailand-260911-free.shp/gis_osm_roads_free_1.shp")%>%
  st_transform(crs = 32647)
```

```{r}
str(roads_raw)
```

```{r}
sort(unique(roads_raw$fclass))
```

```{r}
main_road_classes <- c(
  "motorway",
  "motorway_link",
  "trunk",
  "trunk_link",
  "primary",
  "primary_link",
  "secondary",
  "secondary_link",
  "tertiary",
  "tertiary_link"
)

roads_select <- roads_raw %>%
  filter(fclass %in% main_road_classes)
```

The road networks is now ready to be filtered to the BMR boundary.

```{r}
roads_bmr <- st_intersection(roads_select,st_union(bmr))
```

The boundary and road networks will be plotted for a visual check.

```{r}
tm_shape(bmr) + tm_polygons(col = "grey95") +
  tm_shape(roads_bmr) + tm_lines(col = "steelblue", lwd = 0.5)
```

### 4.3 2022 Traffic Accidents in BMR

```{r}
accident_full <- read_csv("data/aspatial/thai_road_accident_2019_2022.csv")
```

```{r}
head(accident_full)
```

After checking the columns and content, the required data for 2022 in BMR can be filtered out and transformed into an sf.

```{r}
bmr_province_names <- c(
  "Bangkok",
  "Nonthaburi",
  "Pathum Thani",
  "Samut Prakan",
  "Nakhon Pathom",
  "Samut Sakhon"
)

accident_2022 <- accident_full %>%
  filter(
    year(incident_datetime) == 2022,
    province_en %in% bmr_province_names
  )
```

```{r}
accident_sf <- accident_2022 %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 32647)
```

The three dataset will be plotted together for visual check.

```{r}
tm_shape(bmr) + tm_polygons(col = "grey95") +
  tm_shape(roads_bmr) + tm_lines(col = "steelblue", lwd = 0.5) +
  tm_shape(accident_sf) + tm_dots(fill = "red", col = "red", size = 0.02)
```

An interactive map will allow for a more detailed exploration that the event points fall neatly within the road networks.

```{r}
tmap_mode("view")
tm_shape(bmr) + tm_polygons(col = "grey95", fill_alpha = 0.5) +
  tm_shape(roads_bmr) + tm_lines(col = "steelblue", lwd = 0.5) +
  tm_shape(accident_sf) + tm_dots(fill = "red", col = "red", size = 0.2)
```

```{r}
tmap_mode("plot")
```

### 4.4 Conversion to owin and ppp

The sf data set will be converted to owin and ppp, which are required for SPPA. A check on the number of event points from `accident_sf` is first checked using *nrow()* for comparison with after owin clipping.

```{r}
nrow(accident_sf)
```

```{r}
bmr_owin <- as.owin(bmr)

accident_ppp <- as.ppp(accident_sf)
accident_ppp_bmr <- accident_ppp[bmr_owin]

accident_ppp_bmr
```

The summary shows a four point drop, from 3,593 to 3,589. This is a result of the accident event point falling outside the `bmr` boundary, even though they were labelled as BMR province in the data set. As the number is small, about 0.1%, no further investigation is required.

*unmark()* is used to ensure an unmarked pattern `accident_ppp_unmarked` for the subsequent analysis.

```{r}
accident_ppp_unmarked <- unmark(accident_ppp_bmr)
```

## 5. First-Order Point Pattern Analysis

### 5.1 Clark-Evans Test

The ppp will first be re-scaled to kilometers to make the statistical output more interpretable. This is particularly relevant for accident events, as using meters may result in an output of less than one accident.

```{r}
accident_ppp_km <- rescale.ppp(accident_ppp_unmarked, 1000, "km")
```

*Method selection & justification.* The Clark-Evans nearest-neighbour test is used as an initial, well-established first pass to check whether accidents deviate from Complete Spatial Randomness (CSR) before committing to more computationally intensive methods. It compares the observed mean nearest-neighbour distance against the distance expected under a homogeneous Poisson process at the same intensity. This is appropriate as an exploratory screening step, but — as flagged earlier in the report's assumptions section — a purely planar CSR null is a deliberately naive baseline here, since accidents are structurally confined to the road network rather than free to occur anywhere in the polygon. This test is therefore read as a contrast against the network-constrained K/G-function result, not as primary evidence.

*Parameters and assumptions.* correction = "none" applies no edge correction, meaning points near the study-area boundary may have their true nearest neighbour lying just outside the window and undetected — a known source of bias that inflates the appearance of clustering (worth stating explicitly as a limitation). alternative = "clustered" tests the one-sided hypothesis R \< 1, chosen because clustering is the ex-ante expectation for accident data rather than dispersion. The test's null model assumes a homogeneous Poisson process — independent events at uniform intensity — an assumption clearly violated here by both the road network's uneven spatial distribution and underlying population/traffic-exposure variation.

```{r}
clarkevans.test(
  accident_ppp_km,
  correction = "none",
  clipregion = NULL,
  alternative = "clustered"
)
```

### 5.2 Clark-Evans Test (Monte Carlo, 99 Simulations)

*Method selection & justification.* This re-runs the same test using Monte Carlo simulation (99 CSR realisations at the same point count) instead of the analytic Z-test's asymptotic approximation, providing an empirical p-value that doesn't rely on large-sample normality assumptions — a useful robustness check on the analytic result above.

*Parameters and assumptions.* nsim = 99 sets the simulation count; the minimum achievable p-value at this setting is 1/(99+1) = 0.01, which is exactly what was obtained — meaning the observed pattern was more extreme than every single simulated CSR realisation. The same CSR/independence assumptions as the analytic test apply to each simulated realisation.

```{r}
clarkevans.test(
  accident_ppp_km,
  correction = "none",
  clipregion = NULL,
  alternative = "clustered",
  method = "MonteCarlo",
  nsim = 99
)
```

### 5.3 Bandwidth Selection for Kernel Density Estimation

*Method selection & justification.* Following Baddeley et al. (2016), four automatic bandwidth selectors were compared before committing to one for the KDE surface — bw.diggle(), bw.CvL(), bw.scott(), and bw.ppl() — since bandwidth choice is exactly the kind of methodological decision the assignment requires to be justified rather than asserted.

*Parameters and assumptions.* bw.diggle() minimises a cross-validation criterion tuned to isolate one dominant, tight cluster against a CSR background — it assumes essentially a single main hotspot. bw.CvL() uses the Cronie & van Lieshout criterion, generally more stable on multi-cluster patterns and less prone to pathological under-smoothing. bw.scott() is a non-adaptive rule-of-thumb based on the pattern's spread in x and y separately (hence two values), not tuned to the actual clustering structure. bw.ppl() uses likelihood cross-validation and is the method Baddeley et al. specifically recommend for patterns dominated by multiple tight local clusters rather than one large one — a better structural match for accident data, which is expected to hotspot at many individual junctions rather than one citywide cluster.

```{r}
bw.diggle(accident_ppp_km)
bw.CvL(accident_ppp_km)
bw.scott(accident_ppp_km)
bw.ppl(accident_ppp_km)
```

## References

Runfola D, Anderson A, Baier H, Crittenden M, Dowker E, Fuhrig S, et al. (2020) geoBoundaries: A global database of political administrative boundaries. PLoS ONE 15(4): e0231866. https://doi.org/10.1371/journal.pone.0231866.

EPSG Geodetic Parameter Dataset — EPSG:24047 and Transformation 1812. https://epsg.io/

geoBoundaries — Open Database of Political Administrative Boundaries. https://www.geoboundaries.org/

OpenStreetMap contributors — gis_osm_roads_free_1 and gis_osm_traffic_free_1, via Geofabrik.

Okabe, A., & Yamada, I. (2001). The K-function method on a network and its computational implementation. Geographical Analysis, 33(3), 271–290.

Kam, T. S. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/

Thai Road Accident 2019–2022 dataset, Kaggle.
