---
title: "01: Study on Road Accidents in Bangkok Greater Metropolitan Area"
author: "Chia Ka Ying"
date: "2026-09-12"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

## 1. Introduction and Study Context

Road traffic accidents are among the most persistent and spatially patterned public-safety problems facing rapidly urbanising and urbanised regions. Unlike many other point-based public-safety events, traffic accidents are constrained to occur on the road network rather than freely across the two-dimensional landscape . This structural feature of the data has direct consequences for which analytical methods are valid.

This report investigates the spatial distribution of road traffic accidents recorded in **2022** across the **Bangkok Metropolitan Region (BMR)**, comprising Bangkok, Nonthaburi, Nakhon Pathom, Pathum Thani, Samut Prakan and Samut Sakhon. First-order and second-order spatial point pattern analysis (SPPA), with a network-constrained treatment justified by the road-bound nature of the event will be used.

## 2. Research Questions and Objectives

**Primary research questions:**

a.  Do road traffic accidents in the BMR occur with uniform intensity across space, or are they concentrated in identifiable areas (first-order property)?

b.  Do accidents exhibit evidence of spatial interaction, i.e., clustering, inhibition or approximate randomness, relative to an appropriate network-constrained null model (second-order property)?

**Objectives:**

a.  Prepare and integrate real-world geospatial and temporal data for point-pattern analysis.
b.  Investigate first-order properties of the accident point pattern across space, and, where the data allow, across time.
c.  Investigate second-order properties using both planar and network-constrained methods.
d.  Critically evaluate the assumptions, methodological choices and limitations of the methods used.
e.  Communicate findings through appropriate geovisualisation.
f.  Translate findings into public-safety planning implications.

## 3. Data and Scope of Study

## 3.1 Datasets

| \# | Dataset | Geometry | Role |
|------------------|------------------|------------------|--------------------|
| 1 | `geoBoundaries-THA-ADM1-all.shp` | Multi-polygon | Define the BMR study region (six provinces) |
| 2 | `thai_road_accident_2019_2022.csv` | Point (event) | Provides 2022 traffic accidents |
| 3 | `gis_osm_roads_free_1.shp` | Line | Provides road networks (main roads) |

## 3.2 Point Event Definition

A point event in this study is a single recorded road traffic accident occurring in 2022 within the BMR, represented by its reported latitude/longitude. The source data records one row per accident and includes information on the type of vehicle, accident details and casualties.

## 3.3 Data-quality handling applied to point events:

To prepare the data sets for analysis, the following data-quality handling will be applied for the point event data set `thai_road_accident_2019_2022.csv` (see section 4.3.4):

a.  **Missing values:** There are 189 records that are missing latitude and longitude data in the  data set, which will be filtered out.

b.  **Erroneous values:** Under the field "number_of_vehicles_involved", there are five records indicating that there are zero vehicles. The details were explored and these appear to be data entry errors as the accident descriptions mention collisions. These five records will be treated as data entry errors but retained, considering there will not be any analysis involving number of vehicles.

## 3.4 Spatial Study Window

The spatial study window is the union of the six BMR provinces (Bangkok, Nonthaburi, Nakhon Pathom, Pathum Thani, Samut Prakan, Samut Sakhon), dissolved into a single polygon. This boundary is used both as the owin for planar spatstat analysis and as the clipping extent for the road network and traffic-infrastructure layers.

## 4. Data Preparation

```{r}
pacman::p_load(sf, terra, spatstat, spNetwork, tmap, tidyverse, dplyr, units, lubridate)
```

### 4.1 Geographical Boundary of BMR

#### 4.1.1 Importing geographical boundary data

Prior to preparing the geospatial analytical workspace, it is important to check the details of the dataset. This will aid in the extraction of the provinces in BMR into a shapefile `bmr`.

```{r}
adm1 <- st_read("data/geospatial/geoBoundaries-THA-ADM1-all/geoBoundaries-THA-ADM1.shp") %>%
  st_transform(crs = 32647)
```

```{r}
str(adm1)
```

```{r}
adm1[["shapeName"]]
```

#### 4.1.2 Filtering to BMR boundary

By obtaining the values in the province field, the BMR provinces can now be filtered to derive the data set `bmr`.

```{r}
bmr_select <- adm1 %>%
  filter(shapeName %in% c(
    "Bangkok",
    "Nonthaburi Province",
    "Pathum Thani Province",
    "Samut Prakan Province",
    "Nakhon Pathom Province",
    "Samut Sakhon Province"
  ))
```

```{r}
bmr <- bmr_select %>%
  st_union() %>%
  st_sf(name = "Bangkok Greater Metropolitan Area", geometry = .) %>%
  st_make_valid()
```

### 4.2 Road Networks in BMR

#### 4.2.1 Importing road network data

Similarly, the road networks data set will be checked in order to correctly filter the required data. For the purpose of this analysis, we will use only main roads.

```{r}
roads_raw <- st_read("data/geospatial/thailand-260911-free.shp/gis_osm_roads_free_1.shp")%>%
  st_transform(crs = 32647)
```

```{r}
str(roads_raw)
```

```{r}
sort(unique(roads_raw$fclass))
```

#### 4.2.2 Filtering to main road classes

```{r}
main_road_classes <- c(
  "motorway",
  "motorway_link",
  "trunk",
  "trunk_link",
  "primary",
  "primary_link",
  "secondary",
  "secondary_link",
  "tertiary",
  "tertiary_link"
)

roads_select <- roads_raw %>%
  filter(fclass %in% main_road_classes)
```

#### 4.2.3 Filtering road network to BMR boundary

The road networks is now ready to be filtered to the BMR boundary using *st_crop()* and *st_bbox()* based on the boundary defined in `bmr`.

```
bmr_combined <- st_union(bmr)

roads_bmr <- st_intersection(roads_select, bmr_combined)
```

```{r}
roads_bmr <- roads_select %>%
  st_crop(st_bbox(bmr)) %>%
  st_intersection(st_union(bmr))
```

The boundary and road networks will be plotted for a visual check.

```{r}
tm_shape(bmr) + tm_polygons(col = "grey95") +
  tm_shape(roads_bmr) + tm_lines(col = "steelblue", lwd = 0.5)
```

### 4.3 2022 Traffic Accidents in BMR

#### 4.3.1 Importing traffic accident data set

The traffic accident data set will now be imported. As the study is confined to 2022 event points, the columns will be checked to facilitate the filtering.

```{r}
accident_full <- read_csv("data/aspatial/thai_road_accident_2019_2022.csv")
```

```{r}
head(accident_full)
```

#### 4.3.2 Filtering accident event points to 2022 and BMR boundary

After checking the columns and content, the required data for 2022 in BMR can be filtered out and transformed into an sf.

```{r}
bmr_province_names <- c(
  "Bangkok",
  "Nonthaburi",
  "Pathum Thani",
  "Samut Prakan",
  "Nakhon Pathom",
  "Samut Sakhon"
)

accident_2022 <- accident_full %>%
  filter(
    year(incident_datetime) == 2022,
    province_en %in% bmr_province_names
  )
```

#### 4.3.4 Data handling for point events

*summary()* was used to check for missing values. It is observed that there are 189 "NA" values in latitude and longtitude. These rows will be dropped. Interestingly, there are accident records that have zero vehicles involved under "number_of_vehicles_involved".

```{r}
summary(accident_2022)
```

Using *View()*, five records with 0 vehicles involved were explored and these appear to be data entry errors as the accident descriptions mention collisions. These five records will be treated as data entry errors but retained, considering there will not be any analysis involving number of vehicles.

```{r}
accident_2022 %>%
  filter(number_of_vehicles_involved == 0) %>%
  View()
```

The data set will be transformed into an sf data set, while filtering out the 189 records that have missing latitude and longitude values.

```{r}
accident_sf <- accident_2022 %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 32647)
```

The three data sets will be plotted together for visual check.

```{r}
tm_shape(bmr) + tm_polygons(col = "grey95") +
  tm_shape(roads_bmr) + tm_lines(col = "steelblue", lwd = 0.5) +
  tm_shape(accident_sf) + tm_dots(fill = "red", col = "red", size = 0.02)
```

An interactive map will allow for a more detailed exploration that the event points fall neatly within the road networks.

::: callout-note
## Note

The interactive map block is set to *eval: FALSE* to minimise computational overhead and reduce report rendering time.
:::

```{r}
#| eval: false
tmap_mode("view")
tm_shape(bmr) + tm_polygons(col = "grey95", fill_alpha = 0.5) +
  tm_shape(roads_bmr) + tm_lines(col = "steelblue", lwd = 0.5) +
  tm_shape(accident_sf) + tm_dots(fill = "red", col = "red", size = 0.2)
```

```{r}
#| eval: false
tmap_mode("plot")
```

### 4.4 Conversion to owin and ppp

The sf data set will be converted to owin and ppp, which are required for SPPA. A check on the number of event points from `accident_sf` is first checked using *nrow()* for comparison with after owin clipping.

```{r}
nrow(accident_sf)
```

```{r}
bmr_owin <- as.owin(bmr)

accident_ppp <- as.ppp(accident_sf)
accident_ppp_bmr <- accident_ppp[bmr_owin]

accident_ppp_bmr
```

The summary shows a four point drop, from 3,593 to 3,589. This is a result of the accident event point falling outside the `bmr` boundary, even though they were labelled as BMR province in the data set. As the number is small, about 0.1%, no further investigation is required.

*unmark()* is used to ensure an unmarked pattern `accident_ppp_unmarked` for the subsequent analysis.

```{r}
accident_ppp_unmarked <- unmark(accident_ppp_bmr)
```

## 5. First-Order Point Pattern Analysis

First-order spatial point pattern analysis for this study proceeds in two stages, reflecting the network-bound nature of road traffic accidents.

First, planar Clark-Evans test against Complete Spatial Randomness (CSR) is conducted, not as the primary evidence of clustering, but as a deliberate baseline because accidents can occur only on roads, which occupy a small fraction of the study area's surface, a planar CSR null model is almost certain to be rejected regardless of whether genuine excess risk concentration exists. Confirming this rejection statistically therefore serves two purposes:

a.  It establishes a quantified reference point against which the network-constrained result can later be contrasted; and
b.  It provides direct evidence, rather than assumption, that a planar framing is inadequate for this data.

Kernel density estimation (KDE) is then applied in two forms for the same reason, i.e., a planar KDE surface is retained as an exploratory baseline, while a network-constrained KDE (NKDE) is treated as the primary first-order method, since it estimates intensity along the network itself rather than across the full 2-D plane. This avoids the systematic bias of interpreting uneven road density as uneven accident risk.

### 5.1 Clark-Evans Test

**Method selection & justification.** The Clark-Evans nearest-neighbour test is used as an initial step to check whether accidents deviate from CSR before committing to more computationally intensive methods. It compares the observed mean nearest-neighbour distance against the distance expected under a homogeneous Poisson process at the same intensity. This is appropriate as an exploratory step. As highlighted earlier, a purely planar CSR null is a naive baseline here, since accidents are structurally confined to the road network. This test is therefore used as a contrast against the network-constrained K/G-function result, not as primary evidence.

**Parameters and assumptions.** The Clark-Evans test measures each accident's distance to its nearest neighbouring accident. The use of the *correction = "none"* argument would mean that for points near the study-area boundary, the true nearest neighbour may lie just outside the window and go undetected. Instead, the test will measure distance to a farther point that is within the boundary instead.

The *alternative = "clustered"* argument tests the one-sided hypothesis R \< 1, which is chosen because accidents are expected to cluster around hazard-prone locations such as intersections. The test's null model assumes a homogeneous Poisson process where every location is equally likely to have an accident. This assumption does not hold here as accidents can occur only on roads. Accident occurrence is further shaped by underlying traffic volume and population density. Both violations mean the test's rejection of CSR should be read as evidence that a planar random-point model is the wrong lens for this data, not as direct proof of genuine excess risk concentration.

The ppp will first be re-scaled to kilometers to make the statistical output more interpretable. This is particularly relevant for accident events, as using meters may result in an output of less than one accident per meter square.

```{r}
accident_ppp_km <- rescale.ppp(accident_ppp_unmarked, 1000, "km")
```

```{r}
clarkevans.test(
  accident_ppp_km,
  correction = "none",
  clipregion = NULL,
  alternative = "clustered"
)
```

**Interpretation.** With an R-value of 0.26385, which is well below 1, it can be interpreted that there is strong clustering of traffic accidents. The p-value of \< 0.01 indicates that the R-value is statistically significant, and thus to reject the null hypothesis of CSR. This is aligned with our expectations, since accidents are constrained to road networks.

### 5.2 Clark-Evans Test (Monte Carlo, 99 Simulations)

*Method selection & justification.* This re-runs the same test using Monte Carlo simulation, by comparing the observed R-value to the results obtained from 99 simulations of CSR conditional on the observed number of points.

*Parameters and assumptions.* Using *nsim = 99*,

sets the simulation count; the minimum achievable p-value at this setting is 1/(99+1) = 0.01, which is exactly what was obtained — meaning the observed pattern was more extreme than every single simulated CSR realisation. The same CSR/independence assumptions as the analytic test apply to each simulated realisation.

```{r}
clarkevans.test(
  accident_ppp_km,
  correction = "none",
  clipregion = NULL,
  alternative = "clustered",
  method = "MonteCarlo",
  nsim = 99
)
```

**Interpretation.** With the same R-value of 0.26385 and p-value of 0.01 (smallest value achieveable by nsim = 99), the same interpretation can be made. As the conclusion is made using a simulation-based method, the findings is made more robust.

### 5.3 Bandwidth Selection for Kernel Density Estimation (KDE)

Bandwidth the most consequential parameter in KDE. Too small and the surface just reproduces individual points (noise, not pattern), while too large and it smooths away real local hotspots into one vague blob. Since *density()* requires an indicated bandwidth, a check will be done on four different functions for computing bandwidth.

The bandwidth will be calculated using the four methods and plotted for visual inspection.

```{r}
bw.diggle(accident_ppp_km)
bw.CvL(accident_ppp_km)
bw.scott(accident_ppp_km)
bw.ppl(accident_ppp_km)
```

```{r}
#| fig-width: 10
#| fig-height: 6

kde_diggle <- density(accident_ppp_km, sigma = bw.diggle, edge = TRUE)
kde_cvl    <- density(accident_ppp_km, sigma = bw.CvL, edge = TRUE)
kde_scott    <- density(accident_ppp_km, sigma = bw.scott, edge = TRUE)
kde_ppl    <- density(accident_ppp_km, sigma = bw.ppl, edge = TRUE)

par(mfrow = c(2, 2))
plot(kde_diggle, main = "bw.diggle")
plot(kde_cvl, main = "bw.CvL")
plot(kde_scott, main = "bw.scott")
plot(kde_ppl, main = "bw.ppl")
```

**Interpretation.** Comparing the four bandwidth selectors visually rather than by their sigma values alone reveals a clearer picture. bw.diggle and bw.ppl (0.023 km and 0.75 km respectively) both produce maps that trace the underlying road network, with narrow bright lines following street shapes rather than distinct hotspot regions. This indicates under-smoothing, where the estimate at each location is driven by only the handful of nearest accident points. bw.CvL and bw.scott (7.06 km and \~4-5 km) instead collapse the pattern into a single broad, diffuse blob covering a large area. This indicates over-smoothing.

None of the bandwidth selector produces a map suited to identifying discrete, actionable hotspots. This demonstrates that a purely planar kernel density approach struggles to find a bandwidth that meaningfully separates "accident risk" from "road network shape" in data that is inherently constrained to the road network.

bw.ppl is selected as the working bandwidth for the exploratory planar KDE baseline as it is the least extreme under-smoother and the closest to providing some accident point details. The output is treated as a baseline for comparison rather than the primary first-order evidence. This finding directly reinforces the report's methodological decision to rely on Network KDE (NKDE) as the primary first-order method, since NKDE smooths intensity along the network itself and is therefore not subject to the same trade-off between tracing the road pattern and diffusing it into a blob.

### 5.4 Network KDE (NKDE)

```{r}
accident_bmr <- accident_sf %>%
  st_filter(bmr, .predicate = st_within)
```

```{r}
roads_bmr <- roads_bmr %>%
  st_cast("MULTILINESTRING") %>%
  st_cast("LINESTRING")

class(st_geometry(roads_bmr))
```

REVIEW BELOW CODE

```
lixels_bmr <- lixelize_lines(roads_bmr, 200, mindist = 100)

samples_bmr <- lines_center(lixels_bmr)

nkde_bmr <- nkde(
  lines = roads_bmr,
  events = accident_bmr,
  w = rep(1, nrow(accident_bmr)),
  samples = samples_bmr,
  kernel_name = "quartic",
  bw = 500,
  div = "bw.ppl",
  method = "simple",
  digits = 1,
  tol = 1,
  grid_shape = c(50, 50),
  max_depth = 8,
  agg = 5,
  sparse = TRUE,
  verbose = FALSE
)
```

```
samples_bmr$density <- nkde_bmr
samples_bmr_sf <- st_as_sf(samples_bmr)

tm_shape(bmr) + tm_polygons(col = "grey95") +
  tm_shape(samples_bmr_sf) +
  tm_dots(
    fill = "density",
    fill.scale = tm_scale_intervals(style = "quantile", values = "viridis"),
    fill.legend = tm_legend(title = "NKDE Density"),
    size = 0.05
  ) +
  tm_layout(
    main.title = "Network KDE of Accidents — BMR (bw = 500 m)",
    main.title.size = 1,
    legend.outside = TRUE
  )
```

## References

Runfola D, Anderson A, Baier H, Crittenden M, Dowker E, Fuhrig S, et al. (2020) geoBoundaries: A global database of political administrative boundaries. PLoS ONE 15(4): e0231866. https://doi.org/10.1371/journal.pone.0231866.

EPSG Geodetic Parameter Dataset — EPSG:24047 and Transformation 1812. https://epsg.io/

geoBoundaries — Open Database of Political Administrative Boundaries. https://www.geoboundaries.org/

OpenStreetMap contributors — gis_osm_roads_free_1 and gis_osm_traffic_free_1, via Geofabrik.

Okabe, A., & Yamada, I. (2001). The K-function method on a network and its computational implementation. Geographical Analysis, 33(3), 271–290.

Kam, T. S. R for Geospatial Data Science and Analytics. https://r4gdsa.netlify.app/

Thai Road Accident 2019–2022 dataset, Kaggle.
